---
title: "Categorical Outcome Models"
subtitle: "How and Why Are They Used?"
author: "Prof. Steven F. Koch"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  revealjs:
    theme: solarized
knitr:
  opts_chunk:
    echo: true
    eval: true
    cache: true
    concordance: true
    autodep: true
    comment: "#>"
    dpi: 300
    fig.align: "center"
    out.width: 75%
editor: source
bibliography: 
  - module.bib
  - software-refs.bib
  - data-slides.bib
---


```{r}
#| label: setup
#| include: false

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

#knitr::opts_knit$set(kable.force.latex = TRUE)

# data manip
library(tidyverse)
library(haven)

# tables
library(kableExtra)
library(qwraps2)
  #options(qwraps2_markup = "markdown")
library(stargazer)
library(xtable)
library(gt)
  
# analysis  
library(nnet)
library(AER)
library(MASS)
library(e1071)
library(class)



options(xtable.comment = FALSE)

options(scipen = 99,
        np.messages=FALSE)

set.seed(1234)
num.boot <- 399 
num.splits <- 5000
q1 <- 0.025
q2 <- 0.975
nmulti <- 20

```


```{r}
#| label: somefunctions

ihs <- function(col) { 
                transformed <- log((col) +  (sqrt(col)^2+1));
                return(transformed)
}

CM <- function(cm) {
  factor.values.eval <- colnames(cm)
  CM <- matrix(0,nrow(cm),nrow(cm))
  rownames(CM) <- rownames(cm)
  colnames(CM) <- rownames(cm)
  for(i in 1:ncol(cm)) CM[,(1:nrow(cm))[rownames(cm)==factor.values.eval[i]]] <- cm[,i]
  return(list(CM=CM,CCR=sum(diag(CM))/sum(CM)))
}

only_perc <- function(x,
                   digits = getOption("qwraps2_frmt_digits", 2),
                   na_rm = FALSE,
                   show_denom = "ifNA",
                   show_symbol = TRUE,
                   markup = "latex") #getOption("qwraps2_markup", "latex")) 
                   {
  d <- sum(!is.na(x))
  n <- sum(x, na.rm = na_rm)
  p <- frmt(100*n/d, digits)

  if (show_denom == "never") {
    rtn <- paste0(p,"%")
  } else {
    if (show_denom =="always" | any(is.na(x))) {
      rtn <- paste0(frmt(as.integer(n)), "/", frmt(as.integer(d)), " (", p, "%)")
    } else {
      rtn <- paste0(p,"%")
    }
  }

  if (!show_symbol) {
    rtn <- gsub("%", "", rtn)
  }


  if (markup == "latex") {
    rtn <- gsub("%", "\\\\%", rtn)
  }

  return(rtn)
}
```


# Some Background

## Applied research questions 

1. Does education influence employment prospects?

1. Does the size of the household reduce access to:
    a. food?
    a. healthcare?
    a. clean energy?

1. Which illnesses lead to catastrophic health care expenditure?

1. Does alcohol/drug use/debt influence mental health?

1. Does the age of the predidential candidate affect willingness to vote?

## What is the common denominator?

+ The outcome of interest is categorical:
    + employment prospects: employed, unemployed, non-participant
    + access to food: inadequate, adequate, more than adequate
    + catastrophic health care expenditure: catastrophic, non-catastrophic
    + mental health: depressed, not depressed
    + voting: cast ballot or not

## That means?
  
+ Categorical outcomes are not Gaussian 

+ Errors are synonymous with mis-classification

## How should we think about modelling?

+ Two options:
    1. Assume/Pretend from unobserved continuous variable
    1. Use a prediction algorithm
    1. Sometimes, these are the same
    
+ Train (estimate) the model:
    1. Use all the data, no real evaluation options
    1. Create a training set, evaluate on test set

# Some Data


## The data

1. We will use the LCS data [@lcs_2014_2015]

1. We will focus on food adequacy

1. Our controls will be:
    - Household income
    - Gender of household head
    - Number of children in household
    - Number of adults in household

## Reading it in 

```{r}
#| label: lcs
#| dependson: somefunctions

# LCS data
person <- read_dta(file = "../Data/lcs/lcs-2014-2015-persons-final-v1.dta")
household <- read_dta(file = "../Data/lcs/lcs-2014-2015-households-v1.dta")


# hhsize, number of kids (age < 15) in each household
person1 <- person |>
  group_by(UQNO) |>
  summarise(hhsize = n(), # household size by appearance times of each ID
            kids = sum(Q14AGE < 15), # number of child (age<15)
            adults = hhsize - kids
            ) |>
  mutate(UQNO = as.character(UQNO))

person2 <- person |>
  dplyr::select(UQNO, PERSONNO, Q12SEX, Q13POPGROUP) |>
  filter(PERSONNO == 1,
         Q13POPGROUP == 1) |>  
  transmute(UQNO = as.character(UQNO),
            hhh.dummy = ifelse(Q12SEX==1,1,0),
            hhh.male = as_factor(Q12SEX)) |>
  droplevels() 

# household level variables
house.df <- household |>                
  dplyr::select(
    UQNO, income_inkind, Q221AFOOD) |>
  filter(Q221AFOOD != 9) |>
  transmute(
    UQNO = as.character(UQNO), 
    y = as.numeric(income_inkind),  # 63 observations have 0 values... which(house.df$y==0)
    y.ihs = ihs(y/12),
    adequate = factor(Q221AFOOD, levels = c(1,2,3),
                      labels = c("Below","Adequate","Above"))
    ) |>
  droplevels()
  


#merge above 4 dataframes to get a dataframe LCS2014
LCS2014 <- person1 |>
  full_join(person2, by="UQNO") |>
  full_join(house.df, by="UQNO") |>
  drop_na() |>
  filter(adults > 0,
         adults <= 6,
         kids <= 4)

# random subset
ss <- sample(dim(LCS2014)[1], 3500)
df <- LCS2014[ss,] |>
  drop_na()

# save image
#save(df, file="../data/LCS2014df.Rdata")

# data for binary analysis
bdf <- df |>
  filter(adequate != "Above") |>
  droplevels()

```


## Data summary

```{r}
#| label: dstat
#| dependson: 
#|   - lcs 
#|   - somefunctions
#| results: asis
#| size: tiny
#| echo: false
#| tbl-cap: "Household summary statistics by self-assessed food adequacy."

library(dplyr)

df_sum <- df |>
  group_by(adequate) |>
  summarise(
    male_headed = mean(hhh.male == "Male", na.rm = TRUE),

    kids_mean = mean(kids, na.rm = TRUE),
    kids_sd   = sd(kids, na.rm = TRUE),

    adults_mean = mean(adults, na.rm = TRUE),
    adults_sd   = sd(adults, na.rm = TRUE),

    ihs_mean = mean(y.ihs, na.rm = TRUE),
    ihs_sd   = sd(y.ihs, na.rm = TRUE),

    .groups = "drop"
  )

df_table <- df_sum |>
  transmute(
    adequate,
    `Male headed` = sprintf("%.1f%%", 100 * male_headed),
    `Children: mean (sd)` = sprintf("%.2f (%.2f)", kids_mean, kids_sd),
    `Adults: mean (sd)`   = sprintf("%.2f (%.2f)", adults_mean, adults_sd),
    `IHS: mean (sd)`      = sprintf("%.2f (%.2f)", ihs_mean, ihs_sd)
  ) |>
  tidyr::pivot_longer(-adequate, names_to = "Variable", values_to = "Value") |>
  tidyr::pivot_wider(names_from = adequate, values_from = Value)

gt_table <- gt(df_table) |>
  tab_header(
    title = "Household summary statistics",
    subtitle = "By self-assessed food adequacy"
  ) |>
  cols_align(align = "left", columns = Variable) |>
  cols_align(align = "center", -Variable)

gt_table <- gt_table |>
  tab_row_group(
    label = "Household head gender",
    rows = Variable == "Male headed"
  ) |>
  tab_row_group(
    label = "Household Composition",
    rows = Variable %in% c("Children: mean (sd)", "Adults: mean (sd)")
  ) |>
  tab_row_group(
    label = "Inverse hyperbolic sine of income",
    rows = Variable == "IHS: mean (sd)"
  )

gt_table
```

## Brief discussion of data

+ Data is limited
    - Only households headed by Africans
    - Randomly chosen subset of 3500
+ We will consider binary: below v adequate
+ We will consider multinomial: all levels


## Summary code


```{r}
#| eval: false

library(dplyr)

df_sum <- df |>
  group_by(adequate) |>
  summarise(
    male_headed = mean(hhh.male == "Male", na.rm = TRUE),

    kids_mean = mean(kids, na.rm = TRUE),
    kids_sd   = sd(kids, na.rm = TRUE),

    adults_mean = mean(adults, na.rm = TRUE),
    adults_sd   = sd(adults, na.rm = TRUE),

    ihs_mean = mean(y.ihs, na.rm = TRUE),
    ihs_sd   = sd(y.ihs, na.rm = TRUE),

    .groups = "drop"
  )

df_table <- df_sum |>
  transmute(
    adequate,
    `Male headed` = sprintf("%.1f%%", 100 * male_headed),
    `Children: mean (sd)` = sprintf("%.2f (%.2f)", kids_mean, kids_sd),
    `Adults: mean (sd)`   = sprintf("%.2f (%.2f)", adults_mean, adults_sd),
    `IHS: mean (sd)`      = sprintf("%.2f (%.2f)", ihs_mean, ihs_sd)
  ) |>
  tidyr::pivot_longer(-adequate, names_to = "Variable", values_to = "Value") |>
  tidyr::pivot_wider(names_from = adequate, values_from = Value)

gt_table <- gt(df_table) |>
  tab_header(
    title = "Household summary statistics",
    subtitle = "By self-assessed food adequacy"
  ) |>
  cols_align(align = "left", columns = Variable) |>
  cols_align(align = "center", -Variable)

gt_table <- gt_table |>
  tab_row_group(
    label = "Household head gender",
    rows = Variable == "Male headed"
  ) |>
  tab_row_group(
    label = "Household Composition",
    rows = Variable %in% c("Children: mean (sd)", "Adults: mean (sd)")
  ) |>
  tab_row_group(
    label = "Inverse hyperbolic sine of income",
    rows = Variable == "IHS: mean (sd)"
  )

gt_table
```


# Setup Concepts

## Binary model structure

+ A series of observations $(y_i,X_i)$ 
+ Machine learning thinks of these as training data
+ $X_i$ is observed characteristics and


$$
y_i =  \text{Food adequacy} = \begin{cases}
\text{Below adequate} \\
\text{Adequate}
\end{cases}
$$


## Arbitrary values

+ Often, we impose 'arbitrary' numerical values, $\{0, 1 \}$
+ Should be intuitive


$$
y_i =  \text{Food adequacy} = \begin{cases}
0 & \text{ if Below adequate} \\
1 & \text{ if Adequate}
\end{cases}
$$

## Multinomial ranked responses 

+ Keeping all outcomes
+ One could treat them as ranked
    - bad to good
    - good to bad

$$
y_i = \text{Food adequacy} = \begin{cases}
\text{Below adequate} \\
\text{Adequate} \\
\text{More than adequate}
\end{cases}
$$

## More on ranked alternatives

+ It is often preferred that the spacing is both
    - obvious, and
    - constant
+ Here, it is obvious, maybe not constant 'steps'
+ An example with both?


## Multinomial unranked outcomes 

+ If the ordering is not obvious, it is better to not order them

$$
y_i = \text{Employment Status} = \begin{cases}
\text{Employed} \\
\text{Unemployed} \\
\text{Non-participant in the labour force}
\end{cases}
$$


## Another unranked outcome

$$
y_i = \text{Favourite team} = \begin{cases}
\text{Arsenal} \\
\text{Manchester City} \\
\text{Liverpool} \\
\text{Manchester United}
\end{cases}
$$

# Logit/Probit/LPM

## Main ideas

+ Estimate relationship between $X_i$ and 'success' or 'failure'. 
+ 'Success': $E[y_i|X_i] =$prob$(y_i=1|X_i)$.
+ 'Failure': $E[y_i|X_i] =$prob$(y_i=0|X_i)$.
+ How to model probability?
    + Linearly
    + With a probability density function, i.e., MLE
    + Some other way?

## Assume/pretend continuous

+ Noisy measure success $Z_i\gamma_k+e_{ik}$
+ Outcome determined by larger measure from $k=\{0,1\}$


$$
\begin{split}
\text{prob}(y_i=1|X_i) &= \text{prob}(X_i\gamma_1+e_{i1}>X_i\gamma_0 +e_{i0}) \\
&= \text{prob}(X_i\gamma_1 - X_i\gamma_0>e_{i0}-e_{i1}) \\
&= \text{prob}(e_{i0}-e_{i1}<X_i(\gamma_1-\gamma_0)) \\
&= \text{prob}(u_i<X_i\beta)
\end{split}
$$

## An assumption

+ Make probability distribution assumption
    + If we allow $u$ to be normal, we have a probit
    + If we allow $u$ to be logistic, we have a logit
    + If we allow the probability to be linear, we have an LPM


## Linear Probability Model (LPM)

+ Advantages
    - Easy to estimate
    - Easy to interpret

+ Disadvantages
    - Probability laws may be violated
    - Residual is heteroskedastic (easy to fix, though)
    - Marginal effects potentially unreasonable

## Some simulated data

```{r}
#| label: binary1

y <- rbinom(1000,size=1,prob=.4)
u <- rnorm(1000,0,2)
z <- runif(1000,0,1)
x <- ifelse(y==1,10-5*z+u,1+5*z+u)
#plot(x,jitter(y,amount=0.01),
#     xlab="Control Variable",
#     ylab="Outcome")
#legend("right",
#      pch=1,
#      "Observation")
dframe <- data.frame(y,x)
```

## LPM probability issues?

```{r}
#| label: binary-pic
#| dependson: binary1
#| echo: false
#| fig-cap: "Illustration of Binary Outcome (jittered) Against Control Variable Along with Jump Function Guess and a Smooth Function Guess."

model.lm <- lm(y~x,data=dframe)
mod1.b1 <- coefficients(model.lm)[1]
mod1.b2 <- coefficients(model.lm)[2]
xev <- data.frame(x=seq(min(x),max(x),length=25))
yhat <- predict(model.lm,type="response",newdata=xev)
plot(dframe$x,jitter(dframe$y,amount=0.01),
     ylim=c(-.25,1.25),
     xlab="Control Variable",
     ylab="Outcome")
segments(x0=-2,y0=0,x1=5,y1=0,lwd=4,col=2)
segments(x0=5,y0=1,x1=15,y1=1,lwd=4,col=2)
lines(xev$x,yhat,lwd=4,lty=2,col=3)
legend("right",
      pch=c(1,NA,NA),
      lwd=c(NA,4,4),
      lty=c(NA,1,2),
      col=c(1,2,3),
      c("Observation","Jump Guess","Smooth Guess"))
```

## Code for the illustration


```{r}
#| eval: false

model.lm <- lm(y~x,data=dframe)
mod1.b1 <- coefficients(model.lm)[1]
mod1.b2 <- coefficients(model.lm)[2]
xev <- data.frame(x=seq(min(x),max(x),length=25))
yhat <- predict(model.lm,type="response",newdata=xev)
plot(dframe$x,jitter(dframe$y,amount=0.01),
     ylim=c(-.25,1.25),
     xlab="Control Variable",
     ylab="Outcome")
segments(x0=-2,y0=0,x1=5,y1=0,lwd=4,col=2)
segments(x0=5,y0=1,x1=15,y1=1,lwd=4,col=2)
lines(xev$x,yhat,lwd=4,lty=2,col=3)
legend("right",
      pch=c(1,NA,NA),
      lwd=c(NA,4,4),
      lty=c(NA,1,2),
      col=c(1,2,3),
      c("Observation","Jump Guess","Smooth Guess"))
```


# Maximum Likelihood Estimation (MLE)

## Forcing the Probability to be a Probability

+ Rather than linear probability model
+ Assume a probability density - keep to one control for simplicity
+ logit:

$$
\text{prob}(y_i=1|X_i) = \Lambda(\alpha + X_i\beta) = 
     \frac{\exp(\alpha + X_i\beta)}{1+\exp(\alpha + X_i\beta)}
$$

## Probit model

+ The basic model is:

$$
\text{prob}(y_i=1|X_i) = \Phi(X_i\beta) = \int_{-\infty}^{X_i\beta} \phi(z)dz
$$

+ Technically, rather messy to derive the MLE
+ But, basic ideas are the same
+ Computer packages make this fairly easy, though.
+ Parameters are not the same for logit, probit or LPM...

## Logit derivative note

$$
\text{Given:  } \Lambda(z) = \frac{e^z}{1+e^z} 
$$


$$
\begin{split}
\frac{\partial \Lambda(z)}{\partial z} &= \frac{e^z}{1+e^z} -
\frac{e^{2z}}{(1+e^z)^2} \\
 &= \frac{(1+e^z)e^z}{(1+e^z)^2}- \frac{e^{2z}}{(1+e^z)^2} \\
 &= \frac{e^z}{(1+e^z)^2} = \frac{e^z}{1+e^z}\frac{1}{1+e^z} \\
 &= \Lambda(z)(1-\Lambda(z))
\end{split}
$$

## MLE from Bernoulli

+ Assume a probability mass function for likelihood:
+ Recall, $y_i = 0$ or $y_i = 1$


$$
f(y_i|X_i) = p_i^{y_i} (1-p_i)^{(1-y_i)}
$$


+ Assuming IID and logit yields the likelihood:


$$
\mathcal{L} = \prod_{i=1}^N  \Lambda(\alpha + X_i\beta)^{y_i} \big[1-\Lambda(\alpha + X_i\beta)\big]^{(1-y_i)}
$$

## Thus, the log-likelihood


$$
\begin{aligned}
\ell = \ln \mathcal{L} &= \sum_{i=1}^N y_i \cdot \ln \Lambda(\alpha + X_i\beta) \\ 
 &+ 
   (1-y_i) \cdot \ln \big[1-\Lambda(\alpha + X_i\beta)\big]
   \end{aligned}
$$

+ So, what is it when $y_i=0$?
+ $y_i = 1$
+ Logic? Intuition?

## Logit MLE score equation for $\beta$

+ Maximize the log-likelihood over model parameters

$$
\begin{split}
\frac{\partial \ell}{\partial \beta} &= \sum_{i=1}^N y_i \frac{\partial
  \Lambda / \partial \beta}{ \Lambda (\alpha + X_i\beta)} + (1-y_i) 
   \left[ \frac{- \partial  \Lambda / \partial \beta}{1-\Lambda(\alpha + X_i\beta)} \right] = 0 \\
&= \sum_{i=1}^N y_i X_i \frac{\big[\Lambda(1-\Lambda)\big]}{\Lambda} 
  + (1-y_i)(-X_i) \frac{\big[\Lambda(1-\Lambda)\big]}{1-\Lambda}  = 0 \\
&= \sum_{i=1}^N X_i \big[y_i - \Lambda(\alpha + X_i \beta) \big] = 0
\end{split}
$$

## Logit MLE score equation for $\alpha$


$$
\begin{split}
\frac{\partial \ell}{\partial \alpha} &= \sum_{i=1}^N y_i \frac{\partial
  \Lambda / \partial \alpha}{ \Lambda (\alpha + X_i\beta)} + (1-y_i) 
   \left[ \frac{- \partial  \Lambda / \partial \alpha}{1-\Lambda(\alpha + X_i\beta)} \right] = 0 \\
&= \sum_{i=1}^N y_i \frac{\big[\Lambda(1-\Lambda)\big]}{\Lambda} 
  + (1-y_i) \frac{\big[\Lambda(1-\Lambda)\big]}{1-\Lambda}  = 0 \\
&= \sum_{i=1}^N  \big[ y_i - \Lambda(\alpha + X_i \beta) \big] = 0 \\
\hat{\Lambda} &= \overline{y}
\end{split}
$$

## Some comments

+ Must solve both (all) score equations simultaneously
+ Estimation based on numerical methods
+ Thus, slower than linear model, too.

## Estimates of interest from MLE?

+ What does a parameter tell us?
+ Recall the logit probability 


$$
\text{prob}(y_i=1|X_i) = \Lambda(\alpha + X_i\beta)
$$

+ The marginal effect of $X_i$:


$$
\begin{split}
\frac{\partial \text{prob}(y_i=1|X_i)}{\partial X_i} &= \beta
\frac{\partial \Lambda}{\partial X_i} \\
 &= \beta \Lambda (1-\Lambda) 
\end{split}
$$

## Interpretation

+ From the estimates, we see
    - $\beta$ has the same sign as the marginal effect
    - BUT, IT IS NOT THE MARGINAL EFFECT
    - So, limited value to parameter estimates

+ Why might LPMs be preferred?

## Last parameter estimate thoughts, for now

+ Final comments
    - I am assuming continuity of $X$
    - If not continuous, then the derivative is a "difference"
    - Probit is more complicated math
    - Same qualitative conclusion

+ We will return to this

# Variances 

## Basic MLE Variance

+ Variances
    + Outerproduct of the score
    + Derivative of the score.\footnote{The mathematics is greatly simplified by noting that $E[y_i-F]=0$.}


$$
V[\hat\beta] = \left( \sum_{i=1}^N \frac{[(\partial F/ \partial \hat\beta)^2]
    X_i X_i'}{F(1-F)} \right)^{-1}
$$

## Apply analogy principle

+ Use estimates for actual values (inference is 'simple'):

$$
\hat{\beta} \xrightarrow{d} \mathcal{N}(\beta, \mathcal{I}(\hat{\beta})^{-1})
$$

+ But, inference on parameters is not really that important


## Second derivatives under logit


$$
\begin{split}
\frac{\partial \ell}{\partial \alpha} &=  \sum_{i=1}^N  \big[ y_i - \Lambda(\alpha + X_i \beta) \big] \\
\frac{\partial^2 \ell}{\partial \alpha^2} &= - \sum_{i=1}^N \Lambda(1-\Lambda)
\end{split}
$$



$$
\begin{split}
\frac{\partial \ell}{\partial \beta} &= \sum_{i=1}^N X_i \big[y_i - \Lambda(\alpha + X_i \beta) \big]  \\
\frac{\partial^2 \ell}{\partial \beta^2} &= - \sum_{i=1}^N X_i^2 \cdot \Lambda(1-\Lambda)
\end{split}
$$


## One more derivative and the variance matrix 


$$
\begin{split}
\frac{\partial \ell}{\partial \beta} &= \sum_{i=1}^N X_i \big[y_i - \Lambda(\alpha + X_i \beta) \big]  \\
\frac{\partial^2 \ell}{\partial \beta \partial \alpha} &= - \sum_{i=1}^N X_i \cdot \Lambda(1-\Lambda)
\end{split}
$$

+ Matrix $\Omega = [\alpha \text{ } \beta]'$

$$
V[\hat\Omega] = \begin{bmatrix}
- \sum_{i=1}^N \Lambda(1-\Lambda) &  - \sum_{i=1}^N X_i \cdot \Lambda(1-\Lambda) \\
- \sum_{i=1}^N X_i \cdot \Lambda(1-\Lambda) & - \sum_{i=1}^N X_i^2 \cdot \Lambda(1-\Lambda)
\end{bmatrix}^{-1}
$$



## Delta method

+ Remember, our interest is in the marginal effect
+ So, we need its variance, not the one for the parameters
    + From MLE:  $\sqrt{n}(\Omega - \hat\Omega) \rightarrow \mathcal{N}(0,\Sigma)$
    + Thus

$$
\sqrt{n}\left(g(\Omega) - g\left(\hat\Omega \right) \right) \rightarrow \mathcal{N}\left(0,\nabla g(\Omega)' \cdot \Sigma \cdot \nabla g(\Omega)\right)
$$

+ This is similar to the variance rule for squaring a constant
+ The proof of this relies on Taylor expansions

## Marginal effect variance

+ Given the marginal effect of $X_i$:

$$
\frac{\partial \text{prob}(y_i=1|X_i)}{\partial X_i} = \beta \Lambda (1-\Lambda) 
$$

+ A rather messy derivative

$$
\begin{split}
\frac{\partial^2 \text{prob}(y_i=1|X_i)}{\partial X_i \partial \beta} &= \beta \Lambda (1-\Lambda) \\
 &= \Lambda (1-\Lambda) + X_i \beta \Lambda (1-\Lambda)^2 \\ &- X_i\beta\Lambda^2(1-\Lambda)
 \end{split}
$$

+ Or, we can bootstrap..
+ Depends a bit on what we want to consider

# Other Approaches

## Other distribution functions

+ Normality - probit
+ Complementary log-log: $C(X_i\beta) = 1-\exp(-exp(X_i\beta))$.
+ Or $\beta$ or $\Gamma$ distributions
+ These are not commonly applied


## Discriminant analysis

+ Linear DA
+ Quadratic DA
+ We will discuss this
    
## Estimates from quantiles

+ Maximum Score Estimation [@manski1975]
    + Estimate the median $Q_N(\beta) = \sum_i |y_i-\mathbf{1}(X_i\beta>0)|$ 
    + Quantile regression beyond the median [@koenkerbasset1978]

## Less parametric ideas  
    
+ Semiparametric Methods - do not specify $F$.
    + @kleinspady1993 is an early approach
    + There are others
    
+ Fully Nonparametric Methods 
    + Similar to what we discussed with densities and distributions
    + Also, there are clustering approaches

# Thinking about Prediction, instead

## Correct Classification Ratio

+ Map observed outcomes $Y_i=k$, $j \times 1$ vector $\Upsilon_i$


$$
\Upsilon_{ik} = \begin{cases}
 1 & \text{if } Y_i=k \\
 0 & \text{otherwise}
\end{cases}
$$

+ Map predicted outcomes similarly: $\hat{\Upsilon}_i$


 
$$
\hat{\Upsilon}_{ik} = \begin{cases}
 1 & \text{if } \widehat{p}_{ik} = \max_j \{\widehat{p}_{ij} \} \\
 0 & \text{otherwise}
\end{cases}
$$


## Correct Classification Ratio II

+ Define a classification function 
+ Penalize incorrect predictions


$$
Q_i(\Upsilon,\hat{\Upsilon},n) = \begin{cases}
 1 & \text{if } \Upsilon_i = \hat{\Upsilon}_i \\
 0 & \text{otherwise}
\end{cases}
$$

+ Leads to correct classification ratio (CCR)


$$
CCR = n^{-1} \sum_{i=1}^n Q_i(\Upsilon,\hat{\Upsilon},n) 
$$

+ Develop a simple table
+ Actual against predicted outcomes


$$
CM = \Upsilon' \hat{\Upsilon}
$$


## Bayes' classification: what is it?

1. Focus is a bit more on prediction
1. Still something like before:
    + A series of observations $(y_i,X_i)$ 
    + Machine learning thinks of these as training data
    + But, we keep outcomes qualitative


$$
y_i =  \text{Food adequacy} = \begin{cases}
\text{Below adequate} \\
\text{Adequate}
\end{cases}
$$

## Choose max value

+ With two classes, must exceed 0.5
+ With three classes, exceeding 1/3 is not necessarily enough
+ But can choose the max similar to CCR

$$
\text{prob}(y_i = j | X_i = x_0)
$$

## We generalise this a bit 

+ Still using same basic probability model
+ Consider a specific outcome $k$ from $J$ options
+ Technically, we could drop one outcome as a baseline

$$
\text{prob}(y_i = j | X_i = x_0) = \frac{\pi_k f_k(x)} {\sum_{\ell \in J} \pi_\ell f_\ell(x)}
$$

## How do we find $\pi$ and $f$?

+ $\pi$ can be pulled from the data
+ $f$ is often assumed
+ Note: $f$ is over the control/predictor variables
    + Thus, we might model $f$
    + Or, we might consider a less parametric structure

## $k-$ Nearest Neighbors

1. One way to operationalise this classifier
    + Estimate $\pi$ from subsets of the data
    + $f$ is then about choosing subsets
1. Basic intuition:
    + Assign class based on similar observations
    + This depends on how "wide" we make the comparison group

## Mathematically
    
$$
\text{prob}(y_i = j | X_i = x_0) = \frac{1}{k} \sum_{\ell \in N_k} \mathbf{1}(y_\ell = j)
$$

1. Choose $k$
1. Find $k$ neighbors around an observation
1. We see how they are classified, on average
1. We choose largest average

## But, we need to be careful

1. $k$ is like a bandwidth
    + Estimates are sensitive to size of $k$
    + Need a way to 'optimise'
1. Typically, this comes down to some sort of cross-validation
    + Estimate on training data
    + See how well it 'behaves' on evaluation data
    + Behaviour based on some measure of performance (IMSE, Prediction error)

## LDA 

+ Instead, assume $f$ comes from normal
+ Implicitly, $x$ is quantitative (but does not have to be)
+ Estimate $\pi$ from the training data
+ Again, we might need to think about cross-validation
+ Below, only one $x$, easily generalised...

$$
f_k(x) = \frac{1}{\sigma_k \sqrt{2\pi}} \cdot \exp \left[ -\frac{1}{2\sigma^2_k} \left(x - \mu_k \right)^2 \right]
$$

## Linear discriminant choice

+ Assume $J=2$ and constant variance across classes
+ Take logs of the basic probability relationship
+ Yields the basic linear result
+ Find the largest value of 

$$
\delta_k(x) = x \cdot \frac{\mu_k}{\sigma^2} - \frac{\mu_k^2}{2\sigma^2}+\ln \pi_k
$$

## LDA comments

+ It is linear in $x$, so it is linear discriminant analysis
+ Simply put sample estimates in for population estimates (Analogy principle)
+ One could assume different forms for $f$, but normality is very common
+ Certainly, one can allow for more classes

## Quadratic discriminant analysis

+ Allows for different variances in different classes
+ Variances include squared terms
+ Thus, the discriminator includes a quadratic term in $x$

$$
\delta(x) = -\frac{1}{2} \left(x-\mu_k \right)' \Sigma^{-1}_k \left(x-\mu_k \right) -\frac{1}{2} \ln \left| \Sigma_k \right|  + \ln \pi 
$$

+ The first term includes $x'x$
+ It has squared terms and interaction terms
+ Technically, variances and covariances

## Naïve Bayes'

+ In LDA and QDA, we have a joint normal
+ That means the distributions are 'correlated'
+ Instead, assume independence
    - e.g., a series of independent normals
    - Basically, QDA with diagonal $\Sigma$
+ Thus, we replace $f_k$ with $f_{k1}(x_1) \times f_{k2}(x_2) \times \cdots$

$$
\text{prob}(y_i = j | X_i = x_0) = \frac{\pi_k \times f_{k1}(x_1) \times  \cdots \times f_{kJ}(x_J)} {\sum_{\ell \in J} \pi_\ell \times f_{\ell1}(x_1) \times  \cdots \times f_{\ell J}(x_J)}
$$

## Still need estimates

+ We need $\pi_k$ and densities for each $x$
    - $\pi_k$ as average outcomes in training data
    - $x$ can be estimated from histogram or np density

# Predictive performance

## Pseudo-$R^2$


+ A simple performance measure
+ Like OLS, use the intercept model as baseline


$$
\begin{split}
R^2_{b} &=
1-\frac{ \mathcal{L}_N ( \hat{ \beta} )} {\mathcal{L}_N (\overline{y})} \\
 &= 1 - \frac{ \sum_{i=1}^N y_i \ln \hat{p}_i + (1-y_i) \ln
 (1-\hat{p}_i)} {N [ \overline{y} \ln \overline{y} +(1- \overline{y})
 \ln (1-\overline{y} )] }
\end{split}
$$

## Other Performance Measures

+ Other Comparisons
    + Hosmer-Lemeshow
    + ROC Curve and Classification Matrix or CCR
+ We should maybe pay attention to outliers, as well

## Hosmer-Lemeshow I

+ It is a Pearson $\chi^2$ test


$$
\chi_{G-2}^2 = \sum_{g=1}^G \frac{(\text{observed}_g -
  \text{expected}_g)^2}{\text{expected}_g}
$$

+ Observed successes/failures in the data

## Hosmer-Lemeshow II

+ $G$ subgroups of analysis
    + 2 required, more ok
    + Based on the predicted "probabilities "classes"
+ Expected successes/failures 
    + predicted in the model
    + Assume phat$>$0.5 as threshold
+ *Good model works in subgroups, too*



## ROC and Sensitivity

Receiver operating characteristic (ROC) curve plots binary classification as threshold changes. It gives us some idea of sensitivity of results to arbitrary choice of threshold. The area under the curve tells us the probability that a classifier will rank a randomly chosen success higher than a randomly chosen failure, and is similar in spirit to Mann-Whitney and Wilcoxon rank tests.


## ROC II

+ Classifications depend on the chosen cutoff
+ Analysis of tradeoff between sensitivity and specificity
+ ROC curve shows whether model does better than a "guess"
+ Available in STATA and R, and others
+ Not commonly used in economics, though


## Brier Score

+ Rather than using probabilities to get "classes"
+ USE the probabilities

$$
\text{Brier} = N^{-1} \sum_{i=1}^N \left( \hat{p}_i - o_i  \right)^2
$$

+ Basically, a mean squared error
+ Sort of similar to Hosmer-Lemeshow
+ Argued to be a better measure, but not a "test"
+ Lower is better (like AIC/BIC)

# Binary Example 

## Logit, probit  and LPM parameters

```{r}
#| label: binmodel
#| dependson: lcs
#| results: asis
#| echo: false
#| size: tiny

eval.obs <- sample.int(dim(bdf)[1], 250, replace = FALSE)


bdf <- bdf |> mutate(bin.adeq = as.numeric(adequate)-1)

bdf.eval <- bdf[eval.obs,]
bdf.train <- bdf[-c(eval.obs),]

lpm <- lm(bin.adeq ~ y.ihs + adults + kids + hhh.male,
          data = bdf.train)

logit <- glm(adequate ~ y.ihs + adults + kids + hhh.male,
          data = bdf.train,
          family = "binomial")

probit <- glm(adequate ~ y.ihs + adults + kids + hhh.male,
          data = bdf.train,
          family = binomial(link="probit"))

stargazer(lpm, logit, probit, header=FALSE, 
          model.numbers = FALSE,
          dep.var.labels.include = FALSE,
          omit.stat = "all", type="html")
```

## Estimation and table code


```{r}
#| eval: false

eval.obs <- sample.int(dim(bdf)[1], 250, replace = FALSE)


bdf <- bdf |> mutate(bin.adeq = as.numeric(adequate)-1)

bdf.eval <- bdf[eval.obs,]
bdf.train <- bdf[-c(eval.obs),]

lpm <- lm(bin.adeq ~ y.ihs + adults + kids + hhh.male,
          data = bdf.train)

logit <- glm(adequate ~ y.ihs + adults + kids + hhh.male,
          data = bdf.train,
          family = "binomial")

probit <- glm(adequate ~ y.ihs + adults + kids + hhh.male,
          data = bdf.train,
          family = binomial(link="probit"))

stargazer(lpm, logit, probit, header=FALSE, 
          type = "html",
          model.numbers = FALSE,
          dep.var.labels.include = FALSE,
          omit.stat = "all")
```


## Brief discussion

1. Note: I used a training set of 1647 obs
1. All coefficients are different
1. Most are statistically significant
1. LPM gives average marginal effect
    + probit/logit does not
    + probit/logit give direction

## Model predictions

```{r}
#| label: binpredict
#| dependson:  
#|   - binmodel 
#|   - somefunctions
#| results: asis
#| echo: false
#| size: tiny

lpm.pred <- ifelse(predict(lpm, 
                           newdata = bdf.eval) < 0.5, "Less", "OK")

# how many outside unit interval
outside.function <- function(x,nd){
  z1 <- predict(x,newdata=nd)
  z2 <- ifelse(z1 < 0 | z1 > 1, 1, 0)
  share <- sum(z2)/length(z1)
}

lpm.outside <- outside.function(lpm,bdf.eval)

lpm.tab <- table(bdf.eval$adequate,lpm.pred)
lpm.ccr <- 1 - sum(diag(lpm.tab), na.rm = TRUE)/sum(lpm.tab, na.rm=TRUE)

log.pred <- ifelse(predict(logit, 
                           newdata = bdf.eval, 
                           type = "response") < 0.5,  "Less", "OK")
log.tab <- table(bdf.eval$adequate,log.pred)

prob.pred <- ifelse(predict(probit, 
                            newdata = bdf.eval, 
                            type = "response") < 0.5,  "Less", "OK")
prob.tab <- table(bdf.eval$adequate,prob.pred)

predictions <- cbind(lpm.tab,log.tab,prob.tab)

kable(predictions, align = "rrrrrr",
      caption = "Predictive performance by model") |>
  add_header_above(c(" "=1, "LPM"=2, "logit"=2, "probit"=2), 
                   line=TRUE) 
```

+ All the same for "Below/Less"
+ LPM does better for "Adequate/OK"
+ LPM better overall (171 correct)
+ LPM share outside unit interval - `r round(lpm.outside,3)`

## Prediction code


```{r}
#| eval: false

lpm.pred <- ifelse(predict(lpm, 
                           newdata = bdf.eval) < 0.5, "Below", "Adequate")

# how many outside unit interval
outside.function <- function(x,nd){
  z1 <- predict(x,newdata=nd)
  z2 <- ifelse(z1 < 0 | z1 > 1, 1, 0)
  share <- sum(z2)/length(z1)
}

lpm.outside <- outside.function(lpm,bdf.eval)

lpm.tab <- table(bdf.eval$adequate,lpm.pred)
lpm.ccr <- 1 - CM(lpm.tab)$CCR

log.pred <- ifelse(predict(logit, 
                           newdata = bdf.eval, 
                           type = "response") < 0.5,  "Below", "Adequate")
log.tab <- table(bdf.eval$adequate,log.pred)

prob.pred <- ifelse(predict(probit, 
                            newdata = bdf.eval, 
                            type = "response") < 0.5,  "Below", "Adequate")
prob.tab <- table(bdf.eval$adequate,prob.pred)

predictions <- cbind(lpm.tab,log.tab,prob.tab)

kable(predictions, align = "rrrrrr",
      caption = "Predictive performance by model") |>
  add_header_above(c(" "=1, "LPM"=2, "logit"=2, "probit"=2), 
                   line=TRUE) 
```


## LDA and QDA 

```{r}
#| label: discriminant
#| dependson: binmodel
#| results: asis
#| echo: false
#| size: small
#| tbl-cap: "Predictive performance by model"

lin.lda <- lda(adequate ~ y.ihs + adults + kids + hhh.male,
               data = bdf.train)

lda.tab <- table(bdf.eval$adequate,predict(lin.lda, newdata = bdf.eval)$class )

lin.qda <- qda(adequate ~ y.ihs + adults + kids + hhh.male,
               data = bdf.train)

qda.tab <- table(bdf.eval$adequate,predict(lin.qda, newdata = bdf.eval)$class )

predictions <- cbind(lda.tab,qda.tab)

kable(predictions, align = "rrrr") |>
  add_header_above(c(" "=1, "LDA"=2, "QDA"=2), 
                   line=TRUE) 

```

+ LDA does better for "Adequate" and overall
+ QDA does better for "Below"
+ LPM/Logit still in the lead...
    
## LDA and QDA code

```{r}
#| eval: false

lin.lda <- lda(adequate ~ y.ihs + adults + kids + hhh.male,
               data = bdf.train)

lda.tab <- table(bdf.eval$adequate,predict(lin.lda, newdata = bdf.eval)$class )

lin.qda <- qda(adequate ~ y.ihs + adults + kids + hhh.male,
               data = bdf.train)

qda.tab <- table(bdf.eval$adequate,predict(lin.qda, newdata = bdf.eval)$class )

predictions <- cbind(lda.tab,qda.tab)

kable(predictions, align = "rrrr",
      caption = "Predictive performance by model") |>
  add_header_above(c(" "=1, "LDA"=2, "QDA"=2), 
                   line=TRUE) 

```


## $k-$NN and naïve Bayes'

```{r}
#| label: bayes-knn
#| dependson: binmodel
#| results: asis
#| echo: false
#| size: small
#| tbl-cap: "Predictive performance by model"

bayes.class <- naiveBayes(adequate ~ y.ihs + adults + kids + hhh.male,
               data = bdf.train)

bayes.tab <- table(bdf.eval$adequate,predict(bayes.class, newdata = bdf.eval) )

# knn is certainly not a convenient estimator...
# must separate y from x
# nnet has a conflict with select, I think.

bdf.train.knn <- bdf.train |>
  dplyr::select(y.ihs, kids, adults, hhh.dummy) 

train.cl <- bdf.train |> dplyr::select(adequate)

bdf.eval.knn <- bdf.eval |>
  dplyr::select(y.ihs, kids, adults, hhh.dummy)

knn3.pred <- knn(bdf.train.knn, bdf.eval.knn,
                 cl = train.cl$adequate,
                 k=3)

knn3.tab <- table(bdf.eval$adequate, knn3.pred)

knn5.pred <- knn(bdf.train.knn, bdf.eval.knn,
                 cl = train.cl$adequate,
                 k=5)

knn5.tab <- table(bdf.eval$adequate, knn5.pred)

predictions <- cbind(bayes.tab, knn3.tab, knn5.tab) 

colnames(predictions) <- c("Less","OK","Less","OK","Less","OK")

kable(predictions, align = "rrrrrr") |>
  add_header_above(c(" "=1, "Bayes"=2, "KNN 3"=2, "KNN - 5"=2), 
                   line=TRUE) 

```

+ $k-$NN with 3 neighbours does better for "Below"
+ Bayes caught up with LPM 
+ One can never be sure...

## Bayes and $k-$NN code


```{r}
#| eval: false

bayes.class <- naiveBayes(adequate ~ y.ihs + adults + kids + hhh.male,
               data = bdf.train)

bayes.tab <- table(bdf.eval$adequate,predict(bayes.class, newdata = bdf.eval) )

# knn is certainly not a convenient estimator...
# must separate y from x
# nnet has a conflict with select, I think.

bdf.train.knn <- bdf.train |>
  dplyr::select(y.ihs, kids, adults, hhh.dummy) 

train.cl <- bdf.train |> dplyr::select(adequate)

bdf.eval.knn <- bdf.eval |>
  dplyr::select(y.ihs, kids, adults, hhh.dummy)

knn3.pred <- knn(bdf.train.knn, bdf.eval.knn,
                 cl = train.cl$adequate,
                 k=3)

knn3.tab <- table(bdf.eval$adequate, knn3.pred)

knn5.pred <- knn(bdf.train.knn, bdf.eval.knn,
                 cl = train.cl$adequate,
                 k=5)

knn5.tab <- table(bdf.eval$adequate, knn5.pred)

predictions <- cbind(bayes.tab, knn3.tab, knn5.tab) 

colnames(predictions) <- c("Less","OK","Less","OK","Less","OK")

kable(predictions, align = "rrrrrr",
      caption = "Predictive performance by model") |>
  add_header_above(c(" "=1, "Bayes"=2, "KNN 3"=2, "KNN - 5"=2), 
                   line=TRUE) 

```


## General $k-$NN plot: LPM still better

```{r}
#| label: knn-plot
#| echo: false
#| dependson: 
#|  - bayes-knn
#|  - somefunctions
#|  - binpredict
#| fig-cap: "Nearest neighbor performance relative to LPM"

KNN <- NULL
for (i in 1:12){
  knn.pred <- knn(bdf.train.knn, bdf.eval.knn,
                 cl = train.cl$adequate,
                 k=i)
  knn.tab <- table(bdf.eval$adequate, knn.pred)
  knn.ccr <- 1 - CM(knn.tab)$CCR
  rowval <- cbind(k=i,knn.ccr)
  KNN <- rbind(KNN, rowval)
}

knnmin <- min(KNN[,2])
knnmax <- max(KNN[,2])

# old school plot, could do in ggplot2, as well
plot(c(1,12),c(0.25,0.45),
     type="n",
     xlab = "Nearest Neighbours", 
     ylab = "Error Rate")
lines(KNN[,1],KNN[,2],col="red",lty = 1, lwd = 2)
points(KNN[,1],KNN[,2],col="red", pch = 3)
abline(h=lpm.ccr,col="blue", lty = 2, lwd = 2)
legend("top",
       c("KNN","LPM"),
       col=c("red","blue"),
       pch =c(3,NA),
       lty = c(1,2),
       lwd = c(2,2))

```


## Plot code


```{r}
#| eval: false

lpm.ccr <- 1 - CM(lpm.tab)$CCR

KNN <- NULL
for (i in 1:12){
  knn.pred <- knn(bdf.train.knn, bdf.eval.knn,
                 cl = train.cl$adequate,
                 k=i)
  knn.tab <- table(bdf.eval$adequate, knn.pred)
  knn.ccr <- 1 - CM(knn.tab)$CCR
  rowval <- cbind(k=i,knn.ccr)
  KNN <- rbind(KNN, rowval)
}

knnmin <- min(KNN[,2])
knnmax <- max(KNN[,2])

# old school plot, could do in ggplot2, as well
plot(c(1,12),c(0.2,0.45),
     type="n",
     xlab = "Nearest Neighbours", 
     ylab = "Error Rate")
lines(KNN[,1],KNN[,2],col="red",lty = 1, lwd = 2)
points(KNN[,1],KNN[,2],col="red", pch = 3)
abline(h=lpm.ccr,col="blue", lty = 2, lwd = 2)
legend("top",
       c("KNN","LPM"),
       col=c("red","blue"),
       pch =c(3,NA),
       lty = c(1,2),
       lwd = c(2,2))

```



# References


## References {allowframebreaks}

