---
title: "Applied Distribution Analysis"
subtitle: "EKT 813"
author: "Prof SF Koch"
date: "today"
format: 
  revealjs:
    theme: solarized
knitr:
  opts_chunk:
    echo: true
    eval: true
    cache: true
    concordance: true
    autodep: true
    comment: "#>"
    dpi: 300
    fig.align: "center"
    out.width: 75%
editor: source
bibliography: 
  - module.bib
  - software-refs.bib
  - data-slides.bib
---


```{r setup, include=FALSE}

# Defining text size in various chunks 
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

## libraries - data management
library(tidyverse)
library(skimr)
library(readxl)
library(janitor)
library(haven)

## libraries - tables
library(knitr)
library(kableExtra)
library(stargazer)
library(gt)

## libraries - plots
library(ggthemes)

## libraries - inequality/testing
library(ineq)
library(nortest)
library(RSD)
library(stodom)

## Estimation
library(np)
options(np.messages=FALSE)


set.seed(42) # randomness to get same results
```

## Some references for dominance and application

1.  @davidsonduclos2000
1.  @maasoumiracine2016
1.  @barrettdonald2003
1.  @hadar-russell1969
1.  @whitmore1970

## Some packages and data
    
1.  `nortest` [@nortest]
1.  `stodom`  [@stodom]
1.  `RSD`     [@rsd]
1.  LCS Data [@lcs_2014_2015]

## Today's objectives

1.  Discuss and apply the most common distribution test

1.  Develop stochastic dominance 'theory'

1.  Describe dominance tests

1.  Estimate/calculate 

## How might we address the following questions?

1.  Has economic growth been pro-poor or pro-rich?
1.  Which assets provide the best returns for risk averse individuals?
1.  Are tax changes (health expenditures) progressive or regressive?
1.  Are returns to education constant, similar for all races and both sexes?
1.  Does a mean comparison answer these?

# Kolmogorov-Smirnov Tests

## An illustration code

```{r}
#| eval: false


# Simulate non-normal (skewed) data
x <- rexp(1000, rate = 1)   # exponential (very skewed)

df <- data.frame(x = x)

# Compute mean and sd for normal overlay
mu <- mean(x)
sigma <- sd(x)

ggplot(df, aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "lightblue",
                 color = "black",
                 alpha = 0.6) +
  
  # Kernel density of the actual data
  geom_density(color = "blue", linewidth = 1) +
  
  # Normal density overlay
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sigma),
                color = "red",
                linewidth = 1.2,
                linetype = "dashed") +
  
  labs(subtitle = "Blue = empirical density, Red dashed = normal approximation",
       x = "Value",
       y = "Density",
       caption = "Simulated exponential data") +
  
  theme_solarized()
```

## An illustration


```{r}
#| label: density-plot
#| fig-cap: "Exponential density compared to normal"
#| echo: false


# Simulate non-normal (skewed) data
x <- rexp(1000, rate = 1)   # exponential (very skewed)

df <- data.frame(x = x)

# Compute mean and sd for normal overlay
mu <- mean(x)
sigma <- sd(x)

ggplot(df, aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "lightblue",
                 color = "black",
                 alpha = 0.6) +
  
  # Kernel density of the actual data
  geom_density(color = "blue", linewidth = 1) +
  
  # Normal density overlay
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sigma),
                color = "red",
                linewidth = 1.2,
                linetype = "dashed") +
  
  labs(subtitle = "Blue = empirical density, Red dashed = normal approximation",
       x = "Value",
       y = "Density",
       caption = "Simulated exponential data") +
  
  theme_solarized()
```

## Questions

1.  Does the data look normal?
1.  No: 
    a.  It does not,
    a.  But, it might be if logged
1.  How different from a normal?
1.  How would we test this?
    a.  Intuition: Like a mean test
    a.  But, applied on a larger scale

## The simple hypothesis

1.  Compare one distribution to a *known* distribution
1.  Typical..
    a.  $H_0$: data from some known distribution
    b.  $H_a$: data is not from that distribution
1.  Structure of Test
    a.  Take ECDF of your data and plot
    a.  Specify the "known" distribution and Plot
    a.  Measure biggest gap
    a. Compare to critical values (found in tables...)
    
## Test statistics

1.  Two-tailed
$$
D = \sup_x \left| F_0 (x) - \hat{F}(x) \right|
$$

2.  One-tailed (not so common)

$$
D = \sup_x  F_0 (x) - \hat{F}(x) 
$$


## Previous density as ECDFs (code)

```{r}
#| eval: false

ggplot(df, aes(x = x)) +
  
  # Empirical CDF
  stat_ecdf(geom = "step", color = "blue", linewidth = 1) +
  
  # Normal CDF overlay
  stat_function(fun = pnorm,
                args = list(mean = mu, sd = sigma),
                color = "red",
                linewidth = 1.2,
                linetype = "dashed") +
  
  labs(title = "Empirical CDF vs Normal CDF",
       subtitle = "Blue = ECDF, Red dashed = Normal CDF",
       x = "Value",
       y = "Cumulative Probability") +
  
  theme_solarized()
```

## Previous density as ECDFs 

```{r}
#| echo: false
#| label: ecdf-plot
#| dependson: density-plot
#| fig-cap: "ECDF of same data to 'see' KS test"

ggplot(df, aes(x = x)) +
  
  # Empirical CDF
  stat_ecdf(geom = "step", color = "blue", linewidth = 1) +
  
  # Normal CDF overlay
  stat_function(fun = pnorm,
                args = list(mean = mu, sd = sigma),
                color = "red",
                linewidth = 1.2,
                linetype = "dashed") +
  
  labs(title = "Empirical CDF vs Normal CDF",
       subtitle = "Blue = ECDF, Red dashed = Normal CDF",
       x = "Value",
       y = "Cumulative Probability") +
  
  theme_solarized()
```

## What do we see?

1.  Clearly, the curves do not coincide.
1.  What is the biggest gap?
    a.  Probably right at 0?
    a.  Maybe around 1?
1.  What does the test give?
    a.  `ks.test(x, "pnorm", mean = mu, sd = sigma)`
    a.   $D=$ `r round(ks.test(x, "pnorm", mean = mu, sd = sigma)$statistic, digits = 3)`
    a.  $p=$ `r round(ks.test(x, "pnorm", mean = mu, sd = sigma)$p.value, digits = 3)`
    
## Technicality

1.  We used sample data to estimate $\mu$ and $\sigma$
    a.  This affects the $p-$value
    a.  Lilliefors problem 
    a.  `lillie.test(x)` from `nortest`
    a.  $D=$ `r round(nortest::lillie.test(x)$statistic, digits = 3)`
    a.  $p=$ `r round(nortest::lillie.test(x)$p.value, digits = 3)`
1.  Not really a problem with such a super low $p-$value
1.  More a worry with smaller samples, too

## Two-sample version 

1.  Assume $x$ has sample size $N$
1.  Assume $y$ has sample size $M$

$$
D_{n,m} = \sup_x \left| \hat{F}_{n,x} - \hat{G}_{m,y} \right|
$$

## Critical values

1. The reject criteria
$$
D_{n,m} > c(\alpha)\sqrt{\frac{N+M}{NM}}
$$

2.  For $\alpha = 0.05$, $c(\alpha) = 1.358$ 
3.  Computer packages use...

$$
c(\alpha) = \sqrt{-\ln \left( \frac{\alpha}{2}\right) \cdot \frac{1}{2}}
$$

## Income and expenditure data

1. Read in the data 
1. Plot ECDFs
1. Test equality

## Read data

```{r}
#| label: data

# Get the income, province and expend data, 
# Adding 1 to income, so no more '0' values 
# Probably better ways, we are just being quick
household <- read_dta ( "../Data/lcs/lcs-2014-2015-households-v1.dta") |> 
  clean_names() |>
  select (uqno, 
          income = income_inkind/12, 
          expenditure = expenditure_inkind/12) |>
  mutate(uqno = as.character(uqno),
         log_income = log(income+1),
         log_expenditure = log(expenditure)) |>
  drop_na()

hh_df <- household |>
  pivot_longer(cols = c(log_income, log_expenditure), 
               values_to = "x",
               names_to = "xtype") |>
  mutate(xtype = factor(xtype)) |>
  select(x,xtype)

```

## Income and expenditure ECDF (code)

```{r}
#| eval: false
 
ggplot(hh_df, aes(x = x, color = xtype)) +
  
  # set x limits
  xlim(5,15) +
  
  # Define colours
  scale_color_manual(values = c(
    "log_income" = "blue",
    "log_expenditure" = "red")) + 
  
  # Empirical CDF
  stat_ecdf(geom = "step", linewidth = 1) +
  
  labs(title = "Empirical CDF of Log Income and Expenditure",
       subtitle = "Blue = Income ECDF, Red = Expenditure CDF",
       x = "Value",
       y = "Cumulative Probability") +
  
  theme_solarized()
```

## Income and expenditure ECDF

```{r}
#| echo: false
#| label: ecdf-inc-exp
#| dependson: data
#| fig-cap: "Income and expenditure ECDF for illustration"

 
ggplot(hh_df, aes(x = x, color = xtype)) +
  
  xlim(5,15) +
  
  # Define colours
  scale_color_manual(values = c(
    "log_income" = "blue",
    "log_expenditure" = "red")) + 
  
  # Empirical CDF
  stat_ecdf(geom = "step", linewidth = 1) +
  
  labs(title = "Empirical CDF of Log Income and Expenditure",
       subtitle = "Blue = Income ECDF, Red = Expenditure CDF",
       x = "Value",
       y = "Cumulative Probability") +
  
  theme_solarized()
```

## What do we see?

1.  Certainly, there are differences.
1.  Expenditure starts a bit slower, then overtakes
1.  Largest difference
    a.  Near 8.5?
    a.  Near 11.5?
    
## Test results

1.  `ks.test(household$log_income, household$log_expenditure)`
1.  $D=$ `r round(ks.test(household$log_income, household$log_expenditure)$statistic, digits = 3)`
1.  $p=$ `r round(ks.test(household$log_income, household$log_expenditure)$p.value, digits = 3)`
1.  Difference is two large for them to be the same
1.  Can we say anything more?

# Expected Utility Backdrop

## Preferences or choices (in uncertain situations)

1. Roughly this developed from Expected Utility
1. Primary focus was based on the mean and variance
    a. Higher mean: good
    a. Lower variance: good
1. But, expected utility is a function of all moments
1. Therefore, would also need assumptions about all derivatives of utility

## A more general framework 

1. Utilities $u_a$, $u_b$ over prospects $a$, $b$
1. Not worried if this is under certainty
1. Uncertainty would assume expected utility
1. Not sure how this carries over to Alex's work
1. Assume $u_a<u_b$
1. Consider probabilities $\alpha$ and $\beta$ with $\alpha < \beta$

## Independence axiom

$$
(1-\beta)u_a + \beta u_b > (1-\alpha)u_a + \alpha u_b
$$

1. $u_b$ was preferred
1. Higher probability on $u_b$ preferred, too.

## Extend to $n$ prospects - equality

1. Notational convenience: $u_i>u_j$ if $i>j$
1. Probabilities: $\alpha_i$ and $\beta_i$
1. Assume 0: $\alpha_i=\beta_i$, $\forall i$.


$$
\sum_{i=1}^n \beta_i u_i = \sum_{i=1}^n \alpha_i u_i
$$

## Extend to $n$ prospects - preference

1. Assume 1: $\alpha_j>\beta_j$ and $\alpha_k<\beta_k$
1. Assume 2: $\alpha_i=\beta_i$ $\forall i$, $i\neq j,k$

$$
\sum_{i=1}^n \beta_i u_i > \sum_{i=1}^n \alpha_i u_i
$$

## Rearranging with respect to the probabilities

1. By assumption: $\beta_i \leq \alpha_i$, $i= \{1, 2, \dots, j, \dots, k-1\}$
1. Thus, because of $j$ assumption:

$$
\sum_i^{k-1} \beta_i < \sum_i^{k-1} \alpha_i
$$

## Implications for the remainig probabilities

1. By assumption: $\beta_i=\alpha_i$, $i=\{k+1, k+2, \dots, n\}$
1. Note: $\alpha_j$, $\alpha_k$ have to offset $\beta_j$, $\beta_k$
1. Probabilities must sum to one...

$$
\sum_{i=1}^r \beta_i = \sum_{i=1}^r \alpha_i, r=\{k, k+1, \dots, n \}
$$


## Putting this together

$$
\sum_{i=k}^r \beta_i \leq \sum_i \alpha_i, r=\{1, \dots, n \}
$$

1. Cumulative distribution of preferred prospect never exceeds that of the inferior
 prospect.
1. This is independent of utility functional form ($u'>0$, assumed)
1. @hadar-russell1969
    a. If $P$ stochastically dominates $P'$, $P$ is preferred to $P'$.
    a. If $P$ is preferred to $P'$ for all utility functions, $P$ stochastically dominates $P'$
    a. Extending results of @quirksaposnik1962
1. These have a welfare interpretation [@maasoumiracine2016]



# Stochastic Dominance

## First order stochastic dominance (*FSD*)

1. Assume $u'>0$
1. The probability function $g$ is said to be at least as large as the probability function $f$ in the sense of *first order stochastic dominance* if and only if:

$$
G(x_i) \leq F(x_i) \qquad \forall x_i \in X
$$

## Simulated data

```{r}
#| label: sd1

x <- rnorm(500,10,2)
y <- rnorm(500,5,2)

Fx <- ecdf(x)
Gy <- ecdf(y)

dfx <- tibble(x=x, type="x")
dfy <- tibble(x=y, type = "y")

ecdfxy <- rbind(dfx,dfy) |>
  mutate(type = factor(type))

```


## *FSD* - $F$ below $G$


```{r}
#| echo: false
#| label: sd1a
#| dependson: sd1
#| fig-cap: "ECDF for simulated data and First Order SD"

 
ggplot(ecdfxy, aes(x = x, color = type)) +
  
  #xlim(5,15) +
  
  # Define colours
  scale_color_manual(values = c(
    "x" = "blue",
    "y" = "red")) + 
  
  # Empirical CDF
  stat_ecdf(geom = "step", linewidth = 1) +
  
  labs(title = "Empirical CDF of x and y",
       subtitle = "Blue = ECDF x, Red = CDF y",
       x = "Value",
       y = "Cumulative Probability") +
  
  theme_solarized()
```


## Why $u'>0$?

1.  Increasing utility
    a.  Obvious: $u(y) > u(x)$ if $y>x$
    a.  *FSD*: prob$(y) >$ prob$(x)$, when $y>x$
    a.  Reinforcing!

1.  Better outcomes are more likely
1.  So, anyone that prefers more will be better off with the *FSD* distribution

## Second order stochastic dominance (SSD)


1. Continue with $u'>0$
1. Add further restriction: $u'' \leq 0$
1. Risk aversion incorporated, now. 
1. The probability function $g$ is said to be at least as large as the probability function $f$ in the sense of *second order stochastic dominance* if and only if:

$$
\int_{x_1}^x G(y)dy \leq \int_{x_1}^x  F(y)dy \qquad \forall x_i \in X
$$

## SSD Data

```{r}
#| label: sd2

n <- 1000

A <- rnorm(n, mean = 10, sd = 1)
B <- rnorm(n, mean = 10, sd = 2)

df <- tibble(
  x = c(A, B),
  dist = rep(c("A (less risk)", "B (more risk)"), each = n)
) |>
  mutate(dist = factor(dist))
```


## SSD - $F$ and $G$ cross


```{r}
#| echo: false
#| label: sd2a
#| dependson: sd2
#| fig-cap: "ECDF for simulated data and Second Order SD"

ggplot(df, aes(x = x, color = dist)) +
  stat_ecdf(linewidth = 1.2) +
  labs(title = "Second-Order Stochastic Dominance",
       subtitle = "CDF comparison",
       y = "F(x)", x = "Outcome") +
  theme_solarized()
```

## Approximate area difference (code)

```{r}
#| eval: false

grid <- seq(min(df$x), max(df$x), length.out = 1000)

FA <- ecdf(A)(grid)
FB <- ecdf(B)(grid)

area_diff <- cumsum(FB - FA) * diff(grid)[1]

df_area <- data.frame(
  x = grid,
  area = area_diff
)

ggplot(df_area, aes(x = x, y = area)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "SSD Condition",
       subtitle = "Cumulative area difference (should stay ≥ 0)",
       y = "∫(F_B - F_A)", x = "x") +
  theme_solarized()
```

## Approximate area difference: $A$ *SSD* $B$

```{r}
#| label: approx-gap
#| dependson: sd2
#| fig-cap: "The area difference - always positive"
#| echo: false

grid <- seq(min(df$x), max(df$x), length.out = 1000)

FA <- ecdf(A)(grid)
FB <- ecdf(B)(grid)

area_diff <- cumsum(FB - FA) * diff(grid)[1]

df_area <- data.frame(
  x = grid,
  area = area_diff
)

ggplot(df_area, aes(x = x, y = area)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "SSD Condition",
       subtitle = "Cumulative area difference (should stay ≥ 0)",
       y = "∫(F_B - F_A)", x = "x") +
  theme_solarized()
```


## Why concave $u$?

```{r}
#| label: sd2b
#| dependson: sd2

# A concave utility function
u <- function(x){
  sqrt(sqrt(x))
}

EU_A <- mean(u(A))
EU_B <- mean(u(B))

```

1. Let us look at expected utilities over the distributions
    a.  `r round(EU_A, digits = 3)`
    a.  `r round(EU_B, digits = 3)`
1. Expected utility is higher maps to $A$ *SSD* $B$

## Plot utility (code)

```{r}
#| eval: false

df_u <- data.frame(x = grid)

ggplot(df_u, aes(x = x)) +
  stat_function(fun = function(x) sqrt(sqrt(x)),
                linewidth = 1.2) +
  labs(title = "Concave Utility Function",
       subtitle = "Risk aversion, so prefers less spread",
       y = "u(x)", x = "Outcome") +
  theme_solarized()

```

## Plot utility 

```{r}
#| label: SSD-utility
#| dependson: approx-gap
#| fig-cap: "Utility plot"
#| echo: false

df_u <- data.frame(x = grid)

ggplot(df_u, aes(x = x)) +
  stat_function(fun = function(x) sqrt(sqrt(x)),
                linewidth = 1.2) +
  labs(title = "Concave Utility Function",
       subtitle = "Risk aversion ⇒ prefers less spread",
       y = "u(x)", x = "Outcome") +
  theme_solarized()

```


## Third order stochastic dominance (TSD)

1. @whitmore1970
1. Assume: Risk premium associated with uncertainty becomes smaller with wealth
1. Requires: $u'>0$, $u''\leq 0$ and $u''' \geq 0$.
1. The probability function $g$ is said to be at least as large as the 
probability function $f$ in the sense of \textit{third order stochastic dominance}
if and only if:

$$
\begin{aligned}
\int_a^x \int_b^y (F(z) - G(y))dz dy & \geq 0 \qquad \forall x , \text{ and} \\
\int_a^b (F(y)-G(y)) dy &  \geq 0 \qquad \forall x 
\end{aligned}
$$

# Dominance Testing

## @barrettdonald2003 notation^[$SD^j$ implies $SD^{j+1}$]

$$
\begin{aligned}
\mathcal{J}_1(z;G) &= G(z) \\
\mathcal{J}_2(z;G) &= \int_0^z G(t)dt = \int_0^z \mathcal{J}_1(t;G)dt \\
\mathcal{J}_3(z;G) &= \int_0^z \int_0^t G(s) ds dt = \int_0^z \mathcal{J}_2 (t;G) dt
\end{aligned}
$$


## @davidsonduclos2000

$$
\begin{aligned}
\mathcal{J}_j(z;F) &= \frac{1}{(j-1)!} \int_0^z (z-x)^{j-1} dF(x) \\
 &= \frac{1}{N(j-1)!} \sum_{i=1}^N \mathbf{1}\left(X_i \leq z\right) \cdot (z-X_i)^{j-1} 
\end{aligned}
$$

Note that: $\mathbf{1}\left(X_i \leq z\right)$ is the ECDF

## SD and KS

1. Hypotheses
    a. $H_0^j$: $\mathcal{J}_j(z;G) \leq \mathcal{J}_j(z;F) \qquad \forall z \in [0, \overline{z} ]$
    a. $H_a^j$: $\mathcal{J}_j(z;G) > \mathcal{J}_j(z;F) \qquad \text{for some } z \in [0, \overline{z} ]$
1. Test statistic

$$
\hat{S}_j = \sqrt{\frac{NM}{N+M}} \left[ \sup_z \left( \mathcal{J}_j(z;\hat{G}_M) - \mathcal{J}_j(z;\hat{F}_N)  \right) \right]
$$

## Rejection Rules

1. Reject $H_0$ if $\hat{S}_j > c_j$
1. Seems easy, but it is less so in practice
1. Need critical value, $c_j$
    a. For FSD, not too bad
    b. For SSD, TSD, not so nice

## Basically, a KS-Test

1. We can 'easily' calculate $\hat{S}_1$, since it is a simple difference in cdf. 
1. The resulting critical values from $c_1(\alpha) = \sqrt{ -0.5 \ln \alpha }$
1. Recall: $\mathcal{J}_j(z;\hat{G}_M)$ is simply the ECDF over $y$ data
1. So, this seems relatively simple..
    a. Works in obvious cases
    a. Sometimes has wrong size
    a. This has been addressed in most software
1. However, weighting of data is problematic.

## Income and expenditure data

```{r}
#| label: sd-test
#| dependson: data

pr <- rep(1/dim(household)[1],dim(household)[1])
pair <- createStochasticDominance(
  outcome1 = household$log_expenditure,
  outcome2 = household$log_income,
  pr, pr
)

out1 <- household$log_expenditure
out2 <- household$log_income
```

## FSD Plot, again


```{r}
#| echo: false
#| label: ecdf-inc-exp-plot2
#| dependson: data
#| fig-cap: "Income and expenditure ECDF for FSD illustration"

 
ggplot(hh_df, aes(x = x, color = xtype)) +
  
  xlim(5,15) +
  
  # Define colours
  scale_color_manual(values = c(
    "log_income" = "blue",
    "log_expenditure" = "red")) + 
  
  # Empirical CDF
  stat_ecdf(geom = "step", linewidth = 1) +
  
  labs(title = "Empirical CDF of Log Income and Expenditure",
       subtitle = "Blue = Income ECDF, Red = Expenditure CDF",
       x = "Value",
       y = "Cumulative Probability") +
  
  theme_solarized()
```


## The test results

1.  We have already seen the *FSD* plot
1. Using `RSD` package
    a.  *FSD* result `r RSD::fsd.test(pair)`
    a.  *SSD* result `r RSD::ssd.test(pair)`
1. Using `stodom`` package
    a.  *FSD* result `r stodom::fo_stodom(out1,out2,bins_size = 1,useed = 42)[[1]][1,2]`
    a.  *SSD* result `r stodom::so_stodom(out1,out2, bins_size = 1, useed = 42)[[1]][1,2]`
1.  We will look at *SSD* plot next

## SSD plot

```{r}
#| label: ssd-inc-exp
#| dependson: sd-test
#| fig-cap: "SSD illustration for income and expenditure"

ssd.plot(pair,
         names = c("log expenditure","log income")) +
  xlim(5,15)
```

## Using earlier difference in area

```{r}
#| label: inc-exp-gap
#| dependson: sd-test
#| fig-cap: "The CDF area difference for income and expenditure"
#| echo: false

grid <- seq(min(out1), max(out1), length.out = 1000)

F1 <- ecdf(out1)(grid)
F2 <- ecdf(out2)(grid)

area_diff <- cumsum(F2 - F1) * diff(grid)[1]

df_area <- data.frame(
  x = grid,
  area = area_diff
)

ggplot(df_area, aes(x = x, y = area)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "SSD Condition",
       subtitle = "Cumulative area difference (should stay ≥ 0)",
       y = "∫(F_inc - F_exp)", x = "x") +
  theme_solarized()
```



## References 

