---
title: "Categorical Outcome Models"
subtitle: "How and Why Are They Used?"
author: "Prof. Steven F. Koch"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  beamer:
    theme: Madrid
    colortheme: dolphin
    #fonttheme: structureitalicserif
    toc: true
    incremental: false
    slide-level: 3
    keep-tex: true
bibliography: 
  - module.bib
  - software-refs.bib
  - data-slides.bib
include-in-header: preamble.tex
---



```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(cache=TRUE,
                      echo=FALSE,
                      autodep=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      #dev='tikz',
                      dev='pdf',
                      out.width='60%',
                      out.height='60%',
                      fig.align='center')

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

knitr::opts_knit$set(kable.force.latex = TRUE)

# data manip
library(tidyverse)
library(haven)

# tables
library(kableExtra)
library(qwraps2)
  options(qwraps2_markup = "latex")
library(stargazer)
library(xtable)
  
# analysis  
library(nnet)
library(AER)
library(MASS)
library(e1071)
library(class)



options(xtable.comment = FALSE)

options(scipen = 99,
        np.messages=FALSE)

set.seed(42)
num.boot <- 399 
num.splits <- 5000
q1 <- 0.025
q2 <- 0.975
nmulti <- 20

```


```{r}
#| label: some-functions

ihs <- function(col) { 
                transformed <- log((col) +  (sqrt(col)^2+1));
                return(transformed)
}

CM <- function(cm) {
  factor.values.eval <- colnames(cm)
  CM <- matrix(0,nrow(cm),nrow(cm))
  rownames(CM) <- rownames(cm)
  colnames(CM) <- rownames(cm)
  for(i in 1:ncol(cm)) CM[,(1:nrow(cm))[rownames(cm)==factor.values.eval[i]]] <- cm[,i]
  return(list(CM=CM,CCR=sum(diag(CM))/sum(CM)))
}

only_perc <- function(x,
                   digits = getOption("qwraps2_frmt_digits", 2),
                   na_rm = FALSE,
                   show_denom = "ifNA",
                   show_symbol = TRUE,
                   markup = "latex") #getOption("qwraps2_markup", "latex")) 
                   {
  d <- sum(!is.na(x))
  n <- sum(x, na.rm = na_rm)
  p <- frmt(100*n/d, digits)

  if (show_denom == "never") {
    rtn <- paste0(p,"%")
  } else {
    if (show_denom =="always" | any(is.na(x))) {
      rtn <- paste0(frmt(as.integer(n)), "/", frmt(as.integer(d)), " (", p, "%)")
    } else {
      rtn <- paste0(p,"%")
    }
  }

  if (!show_symbol) {
    rtn <- gsub("%", "", rtn)
  }


  if (markup == "latex") {
    rtn <- gsub("%", "\\\\%", rtn)
  }

  return(rtn)
}
```


# Introduction

## Some Background

### Applied research questions 

1. Does education influence employment prospects?

1. Does the size of the household reduce access to:
    a. food?
    a. healthcare?
    a. clean energy?

1. Which illnesses lead to catastrophic health care expenditure?

1. Does alcohol/drug use/debt influence mental health?

1. Does the age of the predidential candidate affect willingness to vote?

### What is the common denominator?

+ The outcome of interest is categorical:
    + employment prospects: employed, unemployed, non-participant
    + access to food: inadequate, adequate, more than adequate
    + catastrophic health care expenditure: catastrophic, non-catastrophic
    + mental health: depressed, not depressed
    + voting: cast ballot or not
  
+ Categorical outcomes are not Gaussian 

+ Errors are synonymous with mis-classification

### How should we think about modelling?

+ Two options:
    1. Assume/Pretend from unobserved continuous variable
    1. Use a prediction algorithm
    1. Sometimes, these are the same
    
+ Train (estimate) the model:
    1. Use all the data, no real evaluation options
    1. Create a training set, evaluate on test set

## Some Data


```{r}
#| label: lcs
#| dependson: some-functions

# LCS data
person <- read_dta(file = "../Data/lcs/lcs-2014-2015-persons-final-v1.dta")
household <- read_dta(file = "../Data/lcs/lcs-2014-2015-households-v1.dta")


# hhsize, number of kids (age < 15) in each household
person1 <- person %>%
  group_by(UQNO) %>%
  summarise(hhsize = n(), # household size by appearance times of each ID
            kids = sum(Q14AGE < 15), # number of child (age<15)
            adults = hhsize - kids
            ) %>%
  mutate(UQNO = as.character(UQNO))

person2 <- person %>%
  dplyr::select(UQNO, PERSONNO, Q12SEX, Q13POPGROUP) %>%
  filter(PERSONNO == 1,
         Q13POPGROUP == 1) %>%  
  transmute(UQNO = as.character(UQNO),
            hhh.dummy = ifelse(Q12SEX==1,1,0),
            hhh.male = as_factor(Q12SEX)) %>%
  droplevels() 

# household level variables
house.df <- household %>%                
  dplyr::select(
    UQNO, income_inkind, Q221AFOOD) %>%
  filter(Q221AFOOD != 9) %>%
  transmute(
    UQNO = as.character(UQNO), 
    y = as.numeric(income_inkind),  # 63 observations have 0 values... which(house.df$y==0)
    y.ihs = ihs(y/12),
    adequate = factor(Q221AFOOD, levels = c(1,2,3),
                      labels = c("Below","Adequate","Above"))
    ) %>%
  droplevels()
  


#merge above 4 dataframes to get a dataframe LCS2014
LCS2014 <- person1 %>%
  full_join(person2, by="UQNO") %>%
  full_join(house.df, by="UQNO") %>%
  drop_na() %>%
  filter(adults > 0,
         adults <= 6,
         kids <= 4)

## random subset
ss <- sample(dim(LCS2014)[1], 2000)
df <- LCS2014[ss,] 

# save image
#save(df, file="../data/LCS2014df.Rdata")

# data for binary analysis
bdf <- df %>%
  filter(adequate != "Above") %>%
  droplevels()

```


### The data

1. We will use the LCS data [@lcs_2014_2015]

1. We will focus on food adequacy

1. Our controls will be:
    - Household income
    - Gender of household head
    - Number of children in household
    - Number of adults in household

1. Data is limited
    - Only households headed by Africans
    - Randomly chosen subset of 2000
    
### Data summary


```{r}
#| label: dstat
#| dependson: lcs; some-functions
#| results: asis
#| size: tiny

our.sum1 <- list(
  "Household head gemder" = 
    list("Male headed" = ~ only_perc(hhh.male=="Male")),
 "Household Composition" = 
    list("Children: mean (sd)" = ~ qwraps2::mean_sd(kids, denote_sd="paren"),
         "Adults: mean (sd)" = ~ qwraps2::mean_sd(adults, denote_sd="paren")),
 "Inverse hyperbolic sine of income" = 
   list("IHS: mean (sd)" = ~ qwraps2::mean_sd(y.ihs, denote_sd="paren"))
)

our.sum1.adequate <- qwraps2::summary_table(dplyr::group_by(df,adequate), our.sum1)


print(our.sum1.adequate, vline="", align="l|rrr",longtable=FALSE,booktabs=T,linesep="",
      caption="Household summary statistics by self-assessed food adequacy.",
      rtitle= "\\bf{Adequate Food}"
      #cnames = c("\\multicolumn{3}{c}{Food Adequacy}")
      )


```

+ We will consider binary: below v adequate
+ We will consider multinomial: all levels


## Setup Concepts

### Binary model structure

+ A series of observations $(y_i,X_i)$ 
+ Machine learning thinks of these as training data
+ $X_i$ is observed characteristics and


$$
y_i =  \text{Food adequacy} = \begin{cases}
\text{Below adequate} \\
\text{Adequate}
\end{cases}
$$

+ Often, we impose 'arbitrary' numerical values, $\{0, 1 \}$
+ Should be intuitive


$$
y_i =  \text{Food adequacy} = \begin{cases}
0 & \text{ if Below adequate} \\
1 & \text{ if Adequate}
\end{cases}
$$

### Multinomial ranked responses 

+ Keeping all outcomes
+ One could treat them as ranked
    - bad to good
    - good to bad

    

$$
y_i = \text{Food adequacy} = \begin{cases}
\text{Below adequate} \\
\text{Adequate} \\
\text{More than adequate}
\end{cases}
$$

+ It is often preferred that the spacing is both
    - obvious, and
    - constant
+ Here, it is obvious, maybe not constant 'steps'
+ An example with both?


### Multinomial unranked outcomes 

+ If the ordering is not obvious, it is better to not order them

$$
y_i = \text{Employment Status} = \begin{cases}
\text{Employed} \\
\text{Unemployed} \\
\text{Non-participant in the labour force}
\end{cases}
$$


+ Or

$$
y_i = \text{Favourite team} = \begin{cases}
\text{Arsenal} \\
\text{Manchester City} \\
\text{Liverpool} \\
\text{Manchester United}
\end{cases}
$$


# Multinomial Models

## Ordered Outcomes


### The Probabilistic Setting

+ $R_i = \gamma Z_i + u_i$ underpins empirical setup


$$
y_i = \begin{cases}
        0, & \text{If } u_i \leq \kappa_1 - \gamma Z_i \\
        1, & \text{If } \kappa_1-\gamma Z_i < u_i \leq \kappa_2 -
        \gamma Z_i \\
        2, & \text{If } \kappa_2 -\gamma Z_i < u_i
\end{cases}
$$

+ For example


$$
\prob(y_i=2|Z_i) = \prob(u_i>\kappa_2 - \gamma Z_i)
$$

+ Need distributional assumption
+ Empirical modeling depends on it


### Dependent Variable Structure

+ Define $p_{ij} \equiv \prob(y_i=j)$
+ There are $J$ possible alternatives (counted from $0$ to $J-1$)
+ With this, introduce a generalisation
    + Assume $F$ is a cdf
    + $\kappa_0 = -\infty$
    + $\kappa_J=\infty$


$$
\begin{split}
\prob(y_i=j|Z_i) &= \prob\left( \kappa_{j-1} - Z_i' \gamma < u_i \leq
  \kappa_j - Z_i' \gamma \right) \\
 &=  F \left(\kappa_j - Z_i' \gamma \right) - F\left( \kappa_{j-1} - Z_i' \gamma \right)
\end{split}
$$


### Density per Observation: Likelihood Contribution

+ Again, define $p_{ij} \equiv \prob(y_i = j)$
+ Assume $J$ alternatives: counted from $0$ to $J-1$


$$
f(y_i,p) = p_{i0}^{y_{i0}} \times p_{i1}^{y_{i1}} \times \cdots p_{iJ-1}^{y_{iJ-1}}
= \prod_{j=0}^{J-1} p_{ij}^{y_{ij}}
$$

+ $N$ independent observations
+ Possible to allow for dependence, through clusters


$$
\begin{split}
f(y,p) &= \prod_{i=1}^N \left(p_{i0}^{y_{i0}} \times p_{i1}^{y_{i1}} \times
  \cdots p_{iJ-1}^{y_{iJ-1}} \right) \\
f(y,\gamma, \kappa) &= \prod_{i=1}^N \prod_{j=0}^{J-1} \left[ F\left( \kappa_j - Z_i'
     \gamma \right) - F\left( \kappa_{j-1} - Z_i' \gamma \right)
 \right]^{y_{ij}} 
\end{split}
$$


### A picture

```{r}
#| label: ologitF
#| out-height: 60%
#| fig-cap: "Ordered logit probabilities with three groups and two assumed cutoffs at -0.5 and 1.0. Thus, the probability that y=0 is the probablility of being below -0.5, while probability y=2 is the probability of being above 1.0..."

x <- rnorm(1000)
F1 <- ecdf(x)

plot(F1)
abline(v = -0.5, col="red")
abline(v = 0.75, col="blue")

```


### Ordered Model

+ Logit
    + Assume $F(v) = \Lambda(v) = e^v/(1+e^v)$
    + Familiar logit assumption
    + Recall: $\Lambda' = \Lambda(1-\Lambda)$
    + The real complication is that 
        + $\kappa$ affects two outcomes (above/below)
        + $\gamma$ affects all outcomes (slope)
    + Return to picture
+ See suggested problems
+ Ordered probit 
    + Assumes normality
    + Math is just a bit messier


### Simulated ordered model illustrated fits

```{r}
#| label: order1
#| fig-cap: "Data and Analysis"

set.seed(9969)

x <- sort(runif(199,0,150))
u <- rnorm(199,0,60)
ystar <- 300 - 2.5*x + u
y.fac <- factor((ystar>75 & ystar <=150)+2*(ystar>150))
y.num <- (ystar>75 & ystar <=150)+2*(ystar>150)
bin.data <- data.frame(y.fac,x,y.num) 


model.ologit <- polr(y.fac~x,
                     data=bin.data,
                     Hess=TRUE,
                     method="logistic")
ologithat <- fitted(model.ologit,type="response")


plot(range(x),c(-0.25,2.25), 
     type = "n",
     xlab="Linear Index of Independent Variables",
     ylab="Dependent Variable") 
points(x,y.num,pch=2,col=1)
abline(lm(y.num~x),col=2)
lines(x,ologithat[,1],col=3,lwd=2,lty=1)
lines(x,ologithat[,2],col=4,lwd=2,lty=1)
lines(x,ologithat[,3],col=5,lwd=2,lty=1)
legend("topright",pch=c(NA,NA,NA,NA),lwd=c(2,2,2,2),
       col=2:5,lty=c(1,1,1,1),bty="n",
       c("Linear","P(0)","P(1)","P(2)"))
```


## Multinomial Logits

### Dependent Variable Structure

+ Recall $p_{ij} \equiv \prob(y_i=j)$
+ There are $J$ possible alternatives
+ Therefore, introduce generalisation


$$
y_{ij} = \begin{cases}
        1, & \text {If alternative } j \text{ is chosen} \\
        0, & \text {Otherwise}
\end{cases}
$$

+ $y_i$ vector of zeroes, and a single 1


$$
y_i = \begin{pmatrix} 0 & 1 & 0 \end{pmatrix}'
$$

### The Multinomial Density per Observation

+ Above information and only three outcomes
+ The multinomial density 


$$
f(y_i,p) = p_{i1}^{y_{i1}} \times p_{i2}^{y_{i2}} \times
p_{i3}^{y_{i3}} = \prod_{j=1}^3 p_{ij}^{y_{ij}}
$$

+ Extended to additional alternatives ($J$ total)


$$
f(y_i,p) = p_{i0}^{y_{i0}} \times p_{i1}^{y_{i1}} \times \cdots p_{iJ-1}^{y_{iJ-1}}
= \prod_{j=0}^{J-1} p_{ij}^{y_{ij}}
$$


### Likelihood Estimation

+ $N$ observations assumed independent 


$$
f(y,p) = \prod_{i=1}^N \left(p_{i0}^{y_{i0}} \times p_{i1}^{y_{i1}} \times
  \cdots p_{iJ}^{y_{iJ}} \right) = \prod_{i=1}^N \prod_{j=1}^J p_{ij}^{y_{ij}}
$$

+ Formalize log likelihood function


$$
\ell = \ln \mathcal{L} = \sum_{i=1}^N \sum_{j=1}^J y_{ij} \ln p_{ij}
$$

+ Probabilities are functions of variables and parameters

### Concerns: IIA

+ Let us consider $p_{i2}$ and $p_{i3}$
+ Assume categry 1 is the baseline

$$
\begin{aligned}
p_{i2} &= \frac{ \exp (x_i \beta_2)}{1 + \exp (x_i \beta_2) + \exp (x_i \beta_3)} \\
p_{i3} &= \frac{ \exp (x_i \beta_3)}{1 + \exp (x_i \beta_2) + \exp (x_i \beta_3)} 
\end{aligned}
$$

+ Consider the ratio of these two probabilities

$$
\frac{p_{i2}}{p_{i3}} = \frac{ \exp (x_i \beta_2)}{ \exp (x_i \beta_3)}
$$

+ It does not depend at all on the baseline category
+ This is IIA, it is a model assumption

### Simulated unordered model illustrated fits

```{r}
#| label: unorder1
#| fig-cap: "Data and Analysis"

set.seed(9969)

x <- sort(runif(199,0,150))
u <- rnorm(199,0,60)
ystar1 <- 300 - 2.5*x + u
ystar2 <- 50 + 1.35*x + u
y <- factor( (ystar1>=75 & ystar2<=200) + 2*(ystar1<75 & ystar2<=200)
             +3*(ystar1>=75 & ystar2>200) )
bin.data <- data.frame(y,x) 

model.mnl <- multinom(y~x,
                  data=bin.data, 
                  trace = FALSE)
mnlhat <- fitted(model.mnl,type="response")


plot(range(x),c(-0.25,4.25), 
     type = "n",
     xlab="Linear Index of Dependent Variables",
     ylab="Independent Variable") 
points(x,y,pch=2,col=1)
#abline(lm(y.num~x),col=2)
lines(x,mnlhat[,1],col=3,lwd=2,lty=1)
lines(x,mnlhat[,2],col=4,lwd=2,lty=1)
lines(x,mnlhat[,3],col=5,lwd=2,lty=1)
lines(x,mnlhat[,4],col=6,lwd=2,lty=1)
#par(mar=c(5.1,4.1,4.1,10.5), xpd=TRUE)
legend("left", inset=c(-0.17,0),
       pch=c(NA,NA,NA,NA),
       lwd=c(2,2,2,2),
       col=3:6,
       lty=c(1,1,1,1),
       bty="n",
       c("P(1)","P(2)","P(3)","P(4)"))
```


## Multinomial Examples


### Model estimates


```{r}
#| label: multimodel
#| dependson: lcs
#| results: asis
#| size: tiny

df <- df %>% 
  mutate(adeq.order = ordered(adequate,
                              levels = c("Below","Adequate","Above")),
         adeq.num = as.numeric(adeq.order) - 1)

eval.obs <- sample.int(dim(df)[1], 250, replace = FALSE)

df.eval <- df[eval.obs,]
df.train <- df[-c(eval.obs),]

lpm <- lm(adeq.num ~ y.ihs + adults + kids + hhh.male,
          data = df.train)

mlogit <- multinom(adequate ~ y.ihs + adults + kids + hhh.male,
                  data = df.train,
                  trace = FALSE)

ologit <- polr(adeq.order ~ y.ihs + adults + kids + hhh.male,
               data = df.train,
               Hess = TRUE)

stargazer(lpm, ologit, mlogit, header=FALSE, 
          omit.stat = "all")
```



### Quick discussion


+ We do not see the same parameter estimates
+ LPM estimates
    + Basically marginal effects 
    + Low to high outcome, as was modelled
    + However, not relative to probabilities
+ Ordered estimates
    + Directionally appropriate (low to high, as was modelled)
    + But, not directly interpreted for probabilities
+ Unordered estimates
    + Relative to baseline (direction)
    + Again, not directly interpreted for probabilities
+ In practice we should calculate marginal effects (below)

### Model performance (LPM is rather more complicated)

```{r}
#| label: multipredict
#| dependson: multimodel
#| results: asis
#| size: small

olog.pred <- predict(ologit, newdata = df.eval, type = "class") 
olog.tab <- table(df.eval$adequate,olog.pred)

mlog.pred <- predict(mlogit, newdata = bdf.eval, type ="class")
mlog.tab <- table(df.eval$adequate,mlog.pred)

predictions <- cbind(olog.tab,mlog.tab)

kable(predictions, align = "rrrrrr",
      caption = "Predictive performance by model") %>%
  add_header_above(c(" "=1, "Ordered"=3, "Unordered"=3), 
                   line=TRUE) 
```

+ Ordered does better for "Below"
+ Ordered does better for "Adequate"
+ No "Above" predictions
+ Ordered better overall
+ Without "Above", LPM doing better?

### LDA and QDA 

```{r}
#| label: multi-discriminant
#| dependson: multimodel
#| results: asis
#| size: small

lin.lda <- lda(adequate ~ y.ihs + adults + kids + hhh.male,
               data = df.train)

lda.tab <- table(bdf.eval$adequate,predict(lin.lda, newdata = df.eval)$class )

lin.qda <- qda(adequate ~ y.ihs + adults + kids + hhh.male,
               data = df.train)

qda.tab <- table(bdf.eval$adequate,predict(lin.qda, newdata = df.eval)$class )

predictions <- cbind(lda.tab,qda.tab)

kable(predictions, align = "rrrrrr",
      caption = "Predictive performance by model") %>%
  add_header_above(c(" "=1, "LDA"=3, "QDA"=3), 
                   line=TRUE) 

```

+ LDA does better for "Adequate" and overall
+ QDA does better for "Below"
+ LDA slightly better overall
+ Not good with "Above"
+ Both better than logit models
    
### $k-$NN and naïve Bayes'

```{r}
#| label: multi-bayes-knn
#| dependson: multimodel
#| results: asis
#| size: small

bayes.class <- naiveBayes(adequate ~ y.ihs + adults + kids + hhh.male,
               data = df.train)

bayes.tab <- table(df.eval$adequate,predict(bayes.class, newdata = df.eval) )

# knn is certainly not a convenient estimator...
# must separate y from x
# nnet has a conflict with select, I think.

df.train.knn <- df.train %>%
  dplyr::select(y.ihs, kids, adults, hhh.dummy)

train.cl <- df.train %>% dplyr::select(adequate)

df.eval.knn <- df.eval %>%
  dplyr::select(y.ihs, kids, adults, hhh.dummy)

knn3.pred <- knn(df.train.knn, df.eval.knn,
                 cl = train.cl$adequate,
                 k=3)

knn3.tab <- table(df.eval$adequate, knn3.pred)

knn5.pred <- knn(df.train.knn, df.eval.knn,
                 cl = train.cl$adequate,
                 k=5)

knn5.tab <- table(df.eval$adequate, knn5.pred)

predictions <- cbind(bayes.tab, knn3.tab)

kable(predictions, align = "rrrrrr",
      caption = "Predictive performance by model") %>%
  add_header_above(c(" "=1, "Bayes"=3, "KNN 3"=3), 
                   line=TRUE) 

```


+ $k-$NN with 3 neighbours does better for "Below"
+ Bayes does better for "Adequate" and overall
+ Not good with "Above"
+ QDA looking better

### General $k-$NN plot: LPM still better

```{r}
#| label: multi-knn-plot
#| dependson: multi-bayes-knn; multipredict
#| fig-cap: "Nearest neighbor performance relative to Ordered Logit"

ologit.ccr <- 1 - CM(olog.tab)$CCR

KNN <- NULL
for (i in 1:30){
  knn.pred <- knn(df.train.knn, df.eval.knn,
                 cl = train.cl$adequate,
                 k=i)
  knn.tab <- table(df.eval$adequate, knn.pred)
  knn.ccr <- 1 - CM(knn.tab)$CCR
  rowval <- cbind(k=i,knn.ccr)
  KNN <- rbind(KNN, rowval)
}

knnmin <- min(KNN[,2])
knnmax <- max(KNN[,2])

# old school plot, could do in ggplot2, as well
plot(c(1,30),c(0.2,0.65),
     type="n",
     xlab = "Nearest Neighbours", 
     ylab = "Error Rate")
lines(KNN[,1],KNN[,2],col="red",lty = 1, lwd = 2)
points(KNN[,1],KNN[,2],col="red", pch = 3)
abline(h=ologit.ccr,col="blue", lty = 2, lwd = 2)
legend("top",
       c("KNN","Ordered"),
       col=c("red","blue"),
       pch =c(3,NA),
       lty = c(1,2),
       lwd = c(2,2))

```


### Ordered marginal effects: below and adequate

```{r}
#| label: ologit-me
#| dependson: multimodel
#| fig-cap: "Ordered estimated marginal effects for income (ihs)"

ihseval <- seq(min(df$y.ihs),max(df$y.ihs), length = 100)

# Not designed to be overly efficient...
# Using natural variation in the x-data
# But, using it for confidence intervals, here
# There, are other options, including packages

prob.bel <- NULL
prob.ab <- NULL

for (i in 1:100){
  df.eval.loop <- df.eval %>% 
    dplyr::select(-y.ihs) %>%
    mutate(y.ihs = ihseval[i])
  prd <- predict(ologit, newdata = df.eval.loop, type = "probs") 
  prob.bel <- rbind(prob.bel,mean(prd[,1]))
  prob.ab <- rbind(prob.ab, mean(prd[,2]))
}

# Have only calculated probabilities
# Now, need to 'difference' the data to get margins
# Again, not meant to be perfect
margins.below <- prob.bel[-1] - prob.bel[-100]
margins.above <- prob.ab[-1] - prob.ab[-100]

ihs.ave <- (ihseval[-1] + ihseval[-100])/2

plot(c(-0.005,0.005),type = "n",
     xlab = "Range of income",
     ylab = "Change in probability")
lines(ihs.ave,margins.below,lty = 1, col ="red", lwd = 2)
lines(ihs.ave,margins.above, lty = 2, col = "blue", lwd = 2)
legend("top",
       c("Below","Above"),
       col = c("red","blue"),
       lty = c(1,2))


```


### Multinomial marginal effects: below and adequate

```{r}
#| label: mlogit-me
#| dependson: multimodel
#| fig-cap: "Unorderd estimated marginal effects for income (ihs)"

ihseval <- seq(min(df$y.ihs),max(df$y.ihs), length = 100)

# Not designed to be overly efficient...
# Using natural variation in the x-data
# But, using it for confidence intervals, here
# There, are other options, including packages

prob.bel <- NULL
prob.ab <- NULL

for (i in 1:100){
  df.eval.loop <- df.eval %>% 
    dplyr::select(-y.ihs) %>%
    mutate(y.ihs = ihseval[i])
  prd <- predict(mlogit, newdata = df.eval.loop, type = "probs") 
  prob.bel <- rbind(prob.bel,mean(prd[,1]))
  prob.ab <- rbind(prob.ab, mean(prd[,2]))
}

# Have only calculated probabilities
# Now, need to 'difference' the data to get margins
# Again, not meant to be perfect
margins.below <- prob.bel[-1] - prob.bel[-100]
margins.above <- prob.ab[-1] - prob.ab[-100]

ihs.ave <- (ihseval[-1] + ihseval[-100])/2

plot(c(-0.005,0.005),type = "n",
     xlab = "Range of income",
     ylab = "Change in probability")
lines(ihs.ave,margins.below,lty = 1, col ="red", lwd = 2)
lines(ihs.ave,margins.above, lty = 2, col = "blue", lwd = 2)
legend("top",
       c("Below","Above"),
       col = c("red","blue"),
       lty = c(1,2))


```




# References


### References {allowframebreaks}

