---
title: "Probability and Distribution"
subtitle: "EKT 813"
author: "Prof SF Koch"
date: "today"
format: 
  revealjs:
    theme: solarized
knitr:
  opts_chunk:
    echo: true
    eval: true
    cache: true
    concordance: true
    autodep: true
    comment: "#>"
    dpi: 300
    fig.align: "center"
    out.width: 75%
editor: source
bibliography: 
  - module.bib
  - software-refs.bib
  - data-slides.bib
---


```{r setup, include=FALSE}

# Defining text size in various chunks 
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

## libraries - data management
library(tidyverse)
library(skimr)
library(readxl)
library(janitor)
library(haven)

## libraries - tables
library(knitr)
library(kableExtra)
library(stargazer)
library(gt)

## libraries - plots
library(ggthemes)

## Estimation
library(np)
options(np.messages=FALSE)


set.seed(42) # randomness to get same results
```

## Some background

1. Much of the discussion comes from @cunningham2021causal

1. There are many other statistics texts that offer similar information

1. We will mostly make use of standard features of R

1. We will also use many of the packages we have used so far

1. I might add in the np-package [@np-pack]

## Today's objectives

1. Review probability

1. Review Bayes' Law/ Theorem

1. Illustrate probability


# Some Basic Definitions

## Random Processes

1. **Random Process** can be repeated many times, gives different outcomes
    a. Discrete: produces a distinct number of outcomes, often represented by integers, but could be colours, or other qualitative measures.
    a. Continuous: produces infinite number of outcomes

1. **Sample Space** is set of all possible outcomes from RP 


## Independence and Causality

1. **Logical Independence** when there is no reason to think two events should happen together
    a. **Post hoc ergo propter hoc**: After this, therefor that
    a. The temporal ordering of events is not sufficient to define cause!
    a. Please, time series people, remember that..
    
1. **Statistical Independence**
    a. The probability of an event happening, does not depend on another event happening
    a. $Pr(A|B) = Pr(A)$
    a. Next slide
    
## Statistical Dependence Example I

1. Probability of an Ace on a draw from a deck of cards?
    a. The sample space is 52
    a. It is discrete: numbers and suits in a deck
    a. There are only 4 aces, one of each suit
    a. 4/52 = 1/13

## Statistical Dependence Example I

2. Suppose a first draw was an ace, what is the probability of a second ace on a second draw (without replacement)?
    a. Sample space is now only 51
    a. only three aces left
    a. 3/51 for second draw if first was an ace
    a. This probability is different if first was not an ace or we sampled with replacement.
    a. Not independent...
    
## Multiple Events I

1. Suppose we have one die, and we roll a five on a first roll.
    a. What is the probability that we roll a five on a second roll?
    a. This is always with replacement, sides do not come off of a die...
    a. There are six sides, and one side has a five.. so 1/6
    a. Independent...
    
## Multiple Events II

2. What is the probability of probability of rolling to fives in a row?
    a. The first is 1/6
    a. The second is 1/6
    a. Together: 1/6 x 1/6 = 1/36
    a. Why?
  
## Back to Cards I

1. Texas hold'em puts two cards down. 
    a. What is probability of having two Aces down?
    a. We will do this for an example before any up cards
    a. One ace: 4/52, Second ace: 3/51
    a. Both aces: 12/2652 = .0045
    a. What would up cards affect?
  
## Back to Cards II

2. Joint Probabilities
    a. Probability of multiple events happening
    a. It is the multiplication of the marginal probability
    a. $Pr(A,B) = Pr(A) Pr(B)$
    a. Why?

## A Two Dice Example (in class)

1. What is the probability of rolling a 7 with two dice?

1. What is the probability of rolling a 3?

1. Are these the same? Why, Why not?

# Probability and Sets

## Some Concepts and Terms

1. Let $A$ be an event, and $B$ be another event

1. With respect to any two events there are four possibilities
    a. $A$ and $B$: they both occur
    a. ~$A$ and $B$: $A$ does not occur, $B$ does
    a. $A$ and ~$B$: $A$ occurs, but $B$ does not
    a. ~$A$ and ~$B$: they both do not occur
    
1. Draw a Venn Diagram (in class)

## Some Additional Concepts and Maths

1. Universe $\mathbf{U}$: everything possible

1. Thus,
    a. $A + \sim A = U$
    a. $B + \sim B = U$
    a. $A + \sim A = B + \sim B$
    a. $A = B + \sim B - \sim A$
    a. $B = A + \sim A - \sim B$
    
## Yes, More Concepts

1. Union: $A \cup B$
    a. All possible events where $A$ **or** $B$ occur
    a. A set containing every element from $A$ **and** every element from $B$
    a. It includes any 'overlap'
    
1. Intersection: $A \cap B$
    a. Set of events that occur together
    a. Any element has to have been in $A$ **and** $B$
    a. It is only the 'overlap'

## More Concepts and Maths I

1. Two components to identify A...
    a. $A = A \cap B + A \cap \sim B$
    a. The part that intersects with B
    a. The part that intersects with ~B

## More Concepts and Maths II

2. Three objects to identify $A \cap B$
    a. $A \cap B = A \cup B - A \cup \sim B + \sim A \cup B$
    a. The set of $A$ outside of $B$
    a. The set of $B$ outside of $A$
    a. The joint set
    a. I am not entirely happy with this one...

2. Look at Venn Diagrams for each (in class)

## An Example Using These Concepts

*Eskom has struggled to produce enough electricity for some time*

a. If this does not improve, the electricity ministered will likely be reshuffled
a. If it does improve, the minister will likely stay
a. Define $A$ to be about electricity production improvements (less loadshedding)
a. Define $B$ as the minister staying put
a. Assume: $Pr(A) = 0.3$, $Pr(B) = 0.6$ and $Pr(A,B) = 0.2$

## Finding Some Results 1

1. What is the probability of a reshuffle and electricity improving?
    a. This is $Pr(A,\sim B)$
    a. $A = A \cup B + A \cup \sim B$
    a. $A \cup \sim B = A - A \cup B$
    a. $Pr(A \cup \sim B) = Pr(A) - Pr(A \cup B)$
    a. $Pr(A \cup \sim B)$ = 0.3 - 0.2 = 0.1

1. Why did we go from  $A \cup \sim B$ to $Pr(A \cup \sim B)$?

## Finding Some Results 2

3. What is the probability of minister stays and electricity improving?
    a. This is $Pr(A,B)$
    a. Given as 0.2

## Finding Some Results 3

 4. What is probability of minister staying and electricity not improving 
    a. This is $Pr(B,\sim A)$
    a. $B = B \cup A + B \cup \sim A$
    a. $B \cup \sim A = B - B \cup A$
    a. $Pr(B \cup \sim A) = Pr(B) - Pr(A \cup B)$
    a. $Pr(B \cup \sim A)$ = 0.6 - 0.2 = 0.4

## Finding Some Results 4
    
5. What is probability of minister reshuffle and electric not improving
    a. This is $Pr(\sim A, \sim B)$
    a. We know $Pr(U) = 1 = Pr(A,B) + Pr(\sim A,B) + Pr(A, \sim B) + Pr(\sim A, \sim B)$
    a. Law of total probability, four possible combinations, only
    a. $Pr(\sim A, \sim B) = 1 - (Pr(A,B) + Pr(\sim A,B) + Pr(A,\sim B))$
    a. $Pr(\sim A, \sim B)$ = 1- .2 - .4 - 1. = 0.3

## Finding Some Results 5
    
6. Another use for total probability
    a. If $Pr(A) = 0.3$, then $Pr(\sim A) = 0.7$
    a. If $Pr(B) = 0.6$, then $Pr(\sim B) = 0.4$

# Conditionality 

## Probability and Sets

1. Sets tell us something about shares

1. For example, 
    a. The share of the universe made up by $A$ is $Pr(A)$
    a. Technically: $Pr(A) = A/\mathbf{U}$

## Probability and Sets II

1. A different example
    a. The share of $A$ is covered by the subset $A \cap B$?
    a.  $(A \cap B)/A = .2/.3 = 2/3$
    a. But, that is 

$$
Pr(B|A) = \frac{Pr(A,B)}{Pr(A)} = .2/.3 = .67
$$

2. In a world where $A$ has occured, what is the probability $B$ has also occured?

## A Conditional Probability

3. We can also calculate the reverse

$$
Pr(A|B) = \frac{Pr(B,A)}{Pr(B)} = .2/.6 = .33
$$

4. These are conditional probabilities!!

5. Conditioning (direction) mostly matters

## More Conditionals I

1. One for the complement of $A$

$$
Pr(B|\sim A) = \frac{Pr(\sim A,B)}{Pr(\sim A)} = 0.4 / 0.7 = 0.57
$$


2. Another for the complement of B

$$
Pr(A|\sim B) = \frac{Pr(\sim B,A)}{Pr(\sim B)} = 0.1 / 0.4 = 0.25
$$


## More Conditionals II

3. Another for both complements

$$
Pr(\sim A|\sim B) = \frac{Pr(\sim B,\sim A)}{Pr(\sim B)} = 0.3 / 0.4 = 0.75
$$

4. Please, double-check!!

## Conditional Probabilities to Joint Probabilities I

1. We already have $Pr(A|B)$ and $Pr(B|A)$

1. Note, both contain $Pr(A,B)=Pr(B,A)$

1. Joint probability comes from simple manipulation
    a. $Pr(A,B) = Pr(A|B)Pr(B)$
    a. $Pr(B,A) = Pr(B|A)Pr(A)$
    a. So, $Pr(A|B)Pr(B) = Pr(B|A)Pr(A)$
  
1. Thus, Naïve Bayes Rule

$$
Pr(A|B) = \frac{Pr(B|A)Pr(A)}{Pr(B)}
$$

## Additional Maths

1. Recall
    a. $A = A \cup B + A \cup \sim B$
    a. Similarly, $B = B \cup A + B \cup \sim A$
    
1. So, we can substitute out Pr(A) and Pr(B)

$$ 
\begin{aligned}
Pr(A|B) &= \frac{Pr(B|A)Pr(A)}{Pr(B)} \\
 &= \frac{Pr(B|A)Pr(A)}{Pr(B \cup A) + Pr(B \cup \sim A)}\\
 &= \frac{Pr(B|A)Pr(A)}{Pr(B,A) + Pr(B,\sim A)} \\
 &= \frac{Pr(B|A)Pr(A)}{Pr(B|A)Pr(A) + Pr(B|\sim A)Pr(\sim A)}
\end{aligned}
$$

## Bayesian Decomposition

1. From previous...

$$ 
Pr(A|B) = \frac{Pr(B|A)Pr(A)}{Pr(B|A)Pr(A) + Pr(B|\sim A)Pr(\sim A)}
$$

2. Likewise...

$$
Pr(B|A) = \frac{Pr(A|B)Pr(B)}{Pr(A|B)Pr(B) + Pr(A|\sim B)Pr(\sim B)}
$$


## Back to Eskom Example

1. What is probability that electricity levels improved, given that minister stayed?

1. We have all the necessary values from above.

$$
\begin{aligned}
Pr(A|B) &= \frac{Pr(B|A)Pr(A)}{Pr(B|A)Pr(A) + Pr(B|\sim A)Pr(\sim A)} \\
  &= \frac{0.2/0.3 \times 0.3}{0.2/0.3 \times 0.3 + 0.4/0.7 \times 0.7} \\
  &= \frac{0.2}{0.2 + 0.4} \\
  &= 0.2/0.6 = 1/3
\end{aligned}
$$

3. Clearly, you could ask more than a few other similar questions


# Bayes and Monty Hall

##  A game show

a. Choose a door from three $(D_1, D_2, D_3)$
a. One has a great prize, not known
a. Two have "bad prizes", not known
a. Monty shows you a different door 
    i.  One not chosen
    i.  Prizes not shuffled
a. Door shown is never a "great prize"
a. You can reconsider your door
a. Question: Should you?

## Further setup

1. Without loss of generality
    a. Assume your door choice was $D_1$
    a. Assume Monty showed $D_3$

1. Initially, $Pr(D_i) = 1/3$ $\forall$ $i$

1. What is effect of learning about one door?

1. In other words, should you take $D_2$, rather than $D_1$?

## Answering 1

1. Initial information: $Pr(D_1) = 1/3$, $Pr(\sim D_1) = 2/3$
    a. In Bayes, we often call these priors
    a. Prior information or probabilities
    
1. New information: $Pr(A_3) = 0$

1. So, $Pr(D_1) = 1/3$ and $Pr(\sim D_1) = 2/3$

1. BUT, $\sim D_1$ is now just $D_2$

1. Switch!


## Answering with Bayes' Law: Door 1


$$
Pr(D_1|D_3) = \frac{Pr(D_3|D_1)Pr(D_1)}{Pr(D_3)} 
$$

What is $Pr(D_3|D_1)$? 

- Recall, $D_1$ chosen.
- There are only two doors to be revealed: 2 or 3.
- $Pr(D_3|D_1) = 1/2$
- So, $Pr(D_3|D_1)Pr(D_1) = 1/2 \times 1/3 = 1/6$

## Answering with Bayes' Law: Door 2

$$
Pr(D_2|D_3) = \frac{Pr(D_3|D_2)Pr(D_2)}{Pr(D_3)} 
$$

What is $Pr(D_3|D_2)$? 

- There are only two doors to be revealed: 2 or 3.
- After $D_3$ is revealed, prize is not in $D_3$
- $Pr(D_3|D_2) = 1$
- So, $Pr(D_3|D_2)Pr(D_2) = 1 \times 1/3 = 1/3$

## Put together 

Also, 

$$
\begin{aligned}
Pr(D_3) = & Pr(D_3|D_1)Pr(D_1) + Pr(D_3|D_2)Pr(D_2) \\
 =& 1/6 +1/3 = 1/2
\end{aligned}
$$

So, for Door 1 and 2

$$
\begin{aligned}
Pr(D_1|D_3) &= \frac{Pr(D_3|D_1)Pr(D_1)}{Pr(D_3)} = \frac{1/6}{1/2} = 1/3 \\
Pr(D_2|D_3) &= \frac{Pr(D_3|D_2)Pr(D_2)}{Pr(D_3)} = \frac{1/3}{1/2} = 2/3
\end{aligned}
$$


## Bayes Decomp 1

1. Bayes Decomp

$$
\begin{aligned}
Pr(D_{1,2}|D_3) =&  \frac{Pr(D_3|D_{1,2})Pr(D_{1,2})}{Z} \\
 Z=& Pr(D_3|D_1)Pr(D_1) + Pr(D_3|D_2)Pr(D_2) \\   +& Pr(D_3|D_3)Pr(D_3)
\end{aligned}
$$

2. What do we mean by $Pr(D_3|D_3)$? 
    a. Probabililty door 3, which is opened, has a great prize
    a. Not so great, we would think? 0?
    
## Bayes Decomp 2

5. How about $Pr(D_3|D_1)$?
    a. 1 is the door picked, so will not be revealed
    a. After reveal, only two doors left
    a. revealed a door with no prize
    a. Thus, the sample space is two doors
    a. probability is now 1/2

## Bayes Decomp 3

6. How about $Pr(D_3|D_2)$?
    a. Door 1 chosen, can't open
    a. Probability of opening $D_3$ if prize behind $D_2$?
    a. That is one - it is not saying the prize is behind $D_2$!!

## Bayes Decomp 4
    
7. What is $Pr(D_1|D_3)$? 
    a. It is not affected by conditioning?
    a. Does that make sense?

$$
\begin{aligned}
Pr(D_1|D_3) &= \frac{Pr(D_3|D_1)Pr(D_1)}{Z} \\
 &= \frac{1/2 \times 1/3}{1/2 \times 1/3 + 1 \times 1/3 + 0 \times 1/3} \\
 &= 1/3
\end{aligned}
$$

## Bayes Decomp 5

8. What is $Pr(D_3|D_3)$? 
    a. It is 0, as implied
    a. Does that make sense?

## Bayes Decomp 6
    
9. What is $Pr(D_2|D_3)$
    a. It is affected by conditioning
    a. Can also figure out via total probability

$$
\begin{aligned}
Pr(D_2|D_3) &= \frac{Pr(D_3|D_2)Pr(D_2)}{Z} \\
 &= \frac{1 \times 1/3}{1/2 \times 1/3 + 1 \times 1/3 + 0 \times 1/3} \\
 &= 2/3
\end{aligned}
$$

# Beginnings of Estimation

## What do we do with econometrics/statistics?

- We are often interested in a population measure
    - Descriptive statistics: mean, median, variance
    - Economic parameters: elasticity, return
    
- We (essentially) never have all the data
- We estimate from a sample
- We hope:
    - Sample 'matches' the relevant population
    - It is repeatable, can be recreated
    - Estimator is appropriate, given the sample

## Sampling

- Focuses on 
    - 'Matching' the population
    - Being repeatable 
    - So, clear detail and description
    
- There are births, deaths, migrations
- Not exactly random events
- Samples are likely off (by a bit)

## Modelling

- Economic theory sits behind most questions
    - It should suggest outcomes/inputs
    - It is a relationship
    - Often presumed causal
    
- The theory link has weakened over time
    - Data is cheaper to access
    - Computing time is cheaper
    - Easier to 'fish' for relationships
    
## Data mining and machine learning

- Used to be a 'dirty' word 
- Becoming more acceptable
- However, it rarely offers the 'parameters' we seek in economics


## Sample mean and variance


$$
\overline{x} = \sum_{i=1}^N \frac{x_i}{N}
$$

$$
s^2 = \sum_{i=1}^N \frac{(x_i - \overline{x})^2}{N-1}
$$

- Why $N-1$?
- What are degrees of freedom?

## Sample minimum and maximum

$$
\min = x_i : x_i \leq x_j \forall j \in N
$$


$$
\max = x_i : x_i \geq x_j \forall j \in N
$$


## Population median

+ Consider a real-valued random variable $X$
+ Consider its "right-continous" distribution function

$$
F(x) = P(X \leq x)
$$

+ Thus, for the median

$$
F^{-1}(0.5) = \inf \{ x:F(x) \geq 0.5 \}
$$

## Sample median

- All is the same, except we need an estimate of $F$
- Call that esimate $\hat{F}$
- Assume $x$ are i.i.d. 
- The empirical cumulative distribution function is an 
unbiased estimate of the CDF

$$
\hat{F}_n(x) = \frac{1}{n} \sum_{i=1}^n \mathbf{1}(x_i \leq x)
$$

## Die roll distribution function code

```{r}
#| eval: false

rolls <- sample(1:6, 1000, replace = TRUE)

df <- data.frame(rolls = rolls)

true_cdf <- data.frame(
  x = 1:6,
  F = cumsum(rep(1/6, 6))
)

ggplot(df, aes(x = rolls)) +
  stat_ecdf(geom = "step", linewidth = 1) +
  geom_point(data = true_cdf, aes(x = x, y = F), color = "red", size = 2) +
  labs(
    title = "Empirical vs True CDF of Die Rolls",
    x = "Roll outcome",
    y = "F(x)",
    caption = "True CDF as red dots"
  ) +
  theme_solarized()

```


## Die roll distribution function plot

```{r}
#| label: die-roll
#| echo: false
#| fig-cap: "Distribution function for 1000 rolls of a die"

rolls <- sample(1:6, 1000, replace = TRUE)

df <- data.frame(rolls = rolls)

true_cdf <- data.frame(
  x = 1:6,
  F = cumsum(rep(1/6, 6))
)

ggplot(df, aes(x = rolls)) +
  stat_ecdf(geom = "step", linewidth = 1) +
  geom_point(data = true_cdf, aes(x = x, y = F), color = "red", size = 2) +
  labs(
    title = "Empirical vs True CDF of Die Rolls",
    x = "Roll outcome",
    y = "F(x)",
    caption = "True CDF as red dots"
  ) +
  theme_solarized()

```

## Smoothing the CDF

```{r}
#| eval: false

###### nonparametric distribution ####

bwdist <- npudistbw(~rolls,
                    data = df,
                    bwmethod = "cv.cdf",
                    ckerorder=4,
                    nmulti=5)

outdist <- plot(bws=bwdist,
            neval=75,
            perspective=FALSE,
            plot.errors.method="bootstrap",
            plot.errors.boot.num=399,
            plot.errors.type="quantiles",
            plot.behavior="data")

npdist.fit <- outdist$d1$dist
npdist.se <- (outdist$d1$derr)
npdist.Ul <- npdist.fit + npdist.se[,2]
npdist.Ll <- npdist.fit + npdist.se[,1]
xevaldist <- outdist$d1$eval$rolls


Fx <- ecdf(rolls)

plot(range(rolls), c(0,1),
     type = "n",
     xlab="Die roll outcome",
     ylab="CDF of 1000 rolls",
     main="Cumulative Distribution of 1000 die rolls")
lines(npdist.fit~xevaldist,col=2,lwd=2,lty=1) 
  lines(npdist.Ul~xevaldist,col=2,lwd=1,lty=2) 
  lines(npdist.Ll~xevaldist,col=2,lwd=1,lty=2) 
lines(Fx,lwd=2,lty=5,col=5)
legend("bottomright",
       lwd=c(2,1,2,2),
       lty=c(1,2,5,3),
       col=c(2,2,5,3),
       #pch=c(NA,NA,16),
       #pt.cex=1,
       c("NP CDF","NP 95\\% CI","Empirical CDF"))
```

## Nonparametric CDF estimate

```{r}
#| echo: false
#| label: np-cdf
#| dependson: die-roll
#| fig-cap: "Nonparametric CDF estimate compared to Empirical CDF"

df <- data.frame(rolls = rolls)

###### nonparametric distribution ####
bwdist <- npudistbw(~rolls,
                    data = df,
                    bwmethod = "cv.cdf",
                    ckerorder=4,
                    nmulti=5)

outdist <- plot(bws=bwdist,
            neval=75,
            perspective=FALSE,
            plot.errors.method="bootstrap",
            plot.errors.boot.num=399,
            plot.errors.type="quantiles",
            plot.behavior="data")

npdist.fit <- outdist$d1$dist
npdist.se <- (outdist$d1$derr)
npdist.Ul <- npdist.fit + npdist.se[,2]
npdist.Ll <- npdist.fit + npdist.se[,1]
xevaldist <- outdist$d1$eval$rolls


Fx <- ecdf(rolls)

plot(range(rolls), c(0,1),
     type = "n",
     xlab="Die roll outcome",
     ylab="CDF of 1000 rolls",
     main="Cumulative Distribution of 1000 die rolls")
lines(npdist.fit~xevaldist,col=2,lwd=2,lty=1) 
  lines(npdist.Ul~xevaldist,col=2,lwd=1,lty=2) 
  lines(npdist.Ll~xevaldist,col=2,lwd=1,lty=2) 
lines(Fx,lwd=2,lty=5,col=5)
legend("bottomright",
       lwd=c(2,1,2,2),
       lty=c(1,2,5,3),
       col=c(2,2,5,3),
       #pch=c(NA,NA,16),
       #pt.cex=1,
       c("NP CDF","NP 95\\% CI","Empirical CDF"))
```


## Continuous variable distribution function code

```{r}
#| eval: false

ln_income <- rnorm(1000,6,1.8)

df <- data.frame(ln_income = ln_income)

ggplot(df, aes(x = ln_income)) +
  stat_ecdf(geom = "step", linewidth = 1) +
  labs(
    title = "Empirical True CDF of Log Income",
    x = "Log Income",
    y = "F(x)"
  ) +
  theme_solarized()

```


## Continuous variable distribution function plot I

```{r}
#| label: income-draw
#| echo: false
#| fig-cap: "Distribution function for 20 draws of log income"

ln_income <- rnorm(20,6,1.8)

df <- data.frame(ln_income = ln_income)

ggplot(df, aes(x = ln_income)) +
  stat_ecdf(geom = "step", linewidth = 1) +
  labs(
    title = "Empirical True CDF of Log Income",
    x = "Log Income",
    y = "F(x)"
  ) +
  theme_solarized()

```


## Continuous variable distribution function plot II

```{r}
#| label: income-draw-1000
#| echo: false
#| fig-cap: "Distribution function for 1000 draws of log income"

ln_income <- rnorm(1000,6,1.8)

df <- data.frame(ln_income = ln_income)

ggplot(df, aes(x = ln_income)) +
  stat_ecdf(geom = "step", linewidth = 1) +
  labs(
    title = "Empirical True CDF of Log Income",
    x = "Log Income",
    y = "F(x)"
  ) +
  theme_solarized()

```


## Histograms

- We jumped ahead a bit to use distributions
- What is a histogram?
- It is a frequency count across bins
    - Bins must be mutually exclusive
    - Typically, either left or right 'inclusive'

$$
c_b = \sum_{b=1}^B m_{ib}
$$

## Formalising the pieces

- Examples
    - $m_{ib} =  \mathbf{1}(L_b \leq x_i < H_b) \forall i \in N$
    - $m_{ib} =  \mathbf{1}(L_b < x_i \leq H_b) \forall i \in N$
- Bin boundaries
    - $L$ is a lower boundary value
    - $H$ is a higher boundary
    - Importantly, $L_1 \leq \min (X)$ and $H_B \geq \max (X)$

## Other ideas

- One can define bins as equally spaced
- $h$ is a bandwidth that is the length of each bin
- Examples:
    - $k = \left[ (\max(X) - \min(X))/h \right]$
    - $k = log_2 n + 1$
- The choices for $h$ and $k$ are numerous
- Some are 'data-driven'

## Die roll histogram code

```{r}
#| eval: false

# Puts two rolls together for something different
df <- data.frame(rolls = rolls) |>
  mutate(group = case_when(
    rolls <= 2 ~ "1–2",
    rolls <= 4 ~ "3–4",
    TRUE ~ "5–6"
  ))

ggplot(df, aes(x = group)) +
  geom_bar() +
  labs(
    title = "Histogram of Die Rolls (3 Groups)",
    x = "Outcome Group",
    y = "Count"
  ) +
  theme_solarized()

```


## Die roll histogram 

```{r}
#| label: die-roll-group
#| dependson: die-roll
#| echo: false
#| fig-cap: "Histogram for 1000 rolls of a die in 3 groupings"

# Puts two rolls together for something different
df <- data.frame(rolls = rolls) |>
  mutate(group = case_when(
    rolls <= 2 ~ "1–2",
    rolls <= 4 ~ "3–4",
    TRUE ~ "5–6"
  ))

ggplot(df, aes(x = group)) +
  geom_bar() +
  labs(
    title = "Histogram of Die Rolls (3 Groups)",
    x = "Outcome Group",
    y = "Count"
  ) +
  theme_solarized()


```

## Densities

- With discrete data
    - Histograms are densities
    - But should be per outcome (not grouped)
    - And, needs to be in probabilities, not frequencies
    
- With continuous data
    - Histogram is a group density
    - Continuous data would have too many bars 
    - So, need a smooth histogram
    
## Estimating a Continuous Density

- What is a density?
- Mathematically, it is a well-defined object


$$
f(x) = \frac{d}{dx}F(x) 
$$

- We measure $F$ via the empirical CDF, from above

$$
\hat{F}_n(x) = \frac{1}{n} \sum_{i=1}^n \mathbf{1}(x_i \leq x)
$$

## A version of a derivative definition


$$
\hat{f}(x) = \frac{F(x+h)-F(x-h)}{2h}
$$

- This is essentiall a histogram
- Data intervals are $2h = h - (-h)$
- This PDF estimate is intuitive
    - The average number of observations in an interval
    - Normalized by the length of the interval

##  Substitute the Empirical CDF into the PDF 

- Unfortunately, histograms are not smooth

$$
\hat{f}(x) = \frac{1}{2nh} \sum_{i=1}^n \left( \mathbf{1} (x_i \leq x+h) - \mathbf{1} (x_i \leq x-h) \right)
$$

- So, some manipulation


$$
\begin{split}
x_i \leq x+ h &\leftrightarrow 
x_i - x &\leq h \\
\frac{x_i-x}{h} &\leq 1
\end{split}
$$

## Evaluating the weak inequality

- Rewrite with absolute values

$$
\left\lvert \frac{x_i - x}{h} \right\rvert \leq 1 
$$

- Break it apart

$$
\begin{split}
\frac{x_i -x}{h} &\leq 1 \\
\frac{x_i - x}{h} & \geq -1
\end{split}
$$

## Make a naïve assumption

- This matches the expressions above

$$
k\left(\frac{x_i-x}{h} \right) = \begin{cases} \frac{1}{2} & \text{if } | \frac{x_i-x}{h} | \leq 1 \\
 0 & \text{otherwise} \end{cases}
$$

- Substituting from above


$$
\begin{split}
\hat{f}(x) &= \frac{1}{nh} \sum_{i=1}^n \left( -\frac{1}{2} \leq \frac{x_i - x} {h} \leq \frac{1}{2} \right) \\
  &= \frac{1}{nh} \sum_{i=1}^n k\left(\frac{x_i-x}{h} \right)
\end{split}
$$

##  The naïve kernel 

- See the following equation 


$$
\hat{f}(x) = \frac{1}{nh} \sum_{i=1}^n k\left(\frac{x_i-x}{h} \right)
$$

- We can use other kernels
    - Gaussian
    - Epanechnikov
    - Triangular
    
## Underlying required properties 

- Kernels should satisfy these 3 properties
- Symmetry is not really required, but easier to use

$$
\begin{split}
\int k(v)dv &= 1  \qquad \text{a proper probability} \\
k(v) &= k(-v) \qquad \text{symmetric} \\
\int v^2 k(v)dv &= \kappa_2 >0 \qquad \text{an existant variance}
\end{split}
$$




## Die roll density/histogram code

```{r}
#| eval: false


df <- data.frame(rolls = rolls)

ggplot(df, aes(x = rolls, y = after_stat(prop))) +
  geom_bar() +
  labs(
    title = "Density histogram of Die Rolls",
    x = "Outcome Group",
    y = "Proportion"
  ) +
  theme_solarized()

```


## Die roll density/histogram 

```{r}
#| label: die-roll-density
#| dependson: die-roll
#| echo: false
#| fig-cap: "Density histogram for 1000 rolls of a die"


df <- data.frame(rolls = rolls)

ggplot(df, aes(x = rolls, y = after_stat(prop))) +
  geom_bar() +
  labs(
    title = "Density histogram of Die Rolls",
    x = "Outcome Group",
    y = "Proportion"
  ) +
  theme_solarized()

```

## Finally, a histogram and density for the 'income' data

```{r}
#| label: income-density
#| dependson: income-draws
#| echo: false
#| fig-cap: "Density and histogram for 1000 draws of a log income"

options(np.messages=FALSE)

###### nonparametric regression ####
bw <- npudensbw(~ln_income,
                data = df,
                bwmethod = "cv.ls",
                ckerorder=4,
                nmulti=5)

out <- plot(bws=bw,
            neval=75,
            perspective=FALSE,
            plot.errors.method="bootstrap",
            plot.errors.boot.num=399,
            plot.errors.type="quantiles",
            plot.behavior="data")

np.fit <- out$d1$dens
np.se <- (out$d1$derr)
np.Ul <- np.fit + np.se[,2]
np.Ll <- np.fit + np.se[,1]
xeval <- out$d1$eval$ln_income

xseq <- seq(min(ln_income),max(ln_income),0.25)
xnorm <- dnorm(xseq,mean(ln_income),sd(ln_income))

hist(ln_income,freq=FALSE,
     ylim=c(0,0.25),
     col="pink",border="red",
     xlab = "The Support of log income",
     ylab = "Density of log income",
     main="Histogram and Density of log income")
lines(np.fit~xeval,col=2,lwd=2,lty=1) 
  lines(np.Ul~xeval,col=2,lwd=1,lty=2) 
  lines(np.Ll~xeval,col=2,lwd=1,lty=2) 
lines(xseq,xnorm,col=4,lwd=2,lty=3)  
legend("topright",
       lwd=c(2,1,2),
       lty=c(1,2,3),
       col=c(2,2,4),
       #pch=c(NA,NA,16),
       #pt.cex=1,
       c("Nonparametric Density","Nonparametric 95\\% CI","Normal Density"))
```

## The code for the above

- It is not in `ggplot2`, but could be
- The data is fairly normal, and the estimate fits the true
- Observations at 1000, so not too surprising


```{r}
#| eval: false

options(np.messages=FALSE)

###### nonparametric regression ####
bw <- npudensbw(~ln_income,
                data = df,
                bwmethod = "cv.ls",
                ckerorder=4,
                nmulti=5)

out <- plot(bws=bw,
            neval=75,
            perspective=FALSE,
            plot.errors.method="bootstrap",
            plot.errors.boot.num=399,
            plot.errors.type="quantiles",
            plot.behavior="data")

np.fit <- out$d1$dens
np.se <- (out$d1$derr)
np.Ul <- np.fit + np.se[,2]
np.Ll <- np.fit + np.se[,1]
xeval <- out$d1$eval$ln_income

xseq <- seq(min(ln_income),max(ln_income),0.25)
xnorm <- dnorm(xseq,mean(ln_income),sd(ln_income))

hist(ln_income,freq=FALSE,
     ylim=c(0,0.25),
     col="pink",border="red",
     xlab = "The Support of log income",
     ylab = "Density of log income",
     main="Histogram and Density of log income")
lines(np.fit~xeval,col=2,lwd=2,lty=1) 
  lines(np.Ul~xeval,col=2,lwd=1,lty=2) 
  lines(np.Ll~xeval,col=2,lwd=1,lty=2) 
lines(xseq,xnorm,col=4,lwd=2,lty=3)  
legend("topright",
       lwd=c(2,1,2),
       lty=c(1,2,3),
       col=c(2,2,4),
       #pch=c(NA,NA,16),
       #pt.cex=1,
       c("Nonparametric Density","Nonparametric 95\\% CI","Normal Density"))
```


## References 

