---
title: "Regression"
subtitle: "Variances, adjustment for testing and decomposition"
author: "Prof. Steven F. Koch"
date: "today"
format: 
  revealjs:
    theme: solarized
knitr:
  opts_chunk:
    echo: true
    eval: true
    cache: true
    concordance: true
    autodep: true
    comment: "#>"
    dpi: 300
    fig.align: "center"
    out.width: 75%
editor: source
bibliography: 
  - module.bib
  - software-refs.bib
  - data-slides.bib
---

## Today's objectives

1. Look more at variances
    a.  Develop heteroskedasticity and cluster fix
    a.  For single parameter tests
    a.  For multiple parameter tests

1. Develop decomposition idea
    a.  Some math
    a.  Some empirics
    
## References

1.  Testing - most any relatively advanced metrics text
    a.  `lmtest` [@lmtest]
    a.  `sandwich` [@sandwich]
    a.  `car` [@car]

1.  Oaxaca-Blinder decomposition
    a.  @oaxaca1973male
    a.  @blinder1973wage
    a.  `oaxaca` [@oaxaca]

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      echo=FALSE,
                      autodep=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      #dev='tikz',
                      dev='pdf',
                      out.width='60%',
                      out.height='60%',
                      fig.align='center')

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

# turn this off for html document
#knitr::opts_knit$set(kable.force.latex = TRUE)

# data manip
library(tidyverse)
library(haven)

# tables
library(kableExtra)
library(qwraps2)
library(stargazer)
library(xtable)
  
# analysis  
library(nnet)
library(AER)
library(MASS)
library(e1071)
library(class)
library(oaxaca)
library(glmnet)

# testing
library(car)
library(sandwich)
library(lmtest)

# special options
#options(scipen = 99,
#        np.messages=FALSE,
#        xtable.comment = FALSE,
#        qwraps2_markup = "latex")

# declaring any special values
set.seed(42)
num.boot <- 399 
num.splits <- 5000
q1 <- 0.025
q2 <- 0.975
nmulti <- 20

```

## Get data from LCS

```{r lcs}
#| label: lcs

ihs <- function(col) { 
                transformed <- log((col) +  (sqrt(col)^2+1))
                return(transformed)
}

# LCS data
person <- read_dta(file = "../Data/lcs/lcs-2014-2015-persons-final-v1.dta")
household <- read_dta(file = "../Data/lcs/lcs-2014-2015-households-v1.dta")


# hhsize, number of kids (age < 15) in each household
person1 <- person |>
  group_by(UQNO) |>
  summarise(hhsize = n(), # household size by appearance times of each ID
            kids = sum(Q14AGE < 15), # number of child (age<15)
            adults = hhsize - kids
            ) |>
  mutate(UQNO = as.character(UQNO))


# household level variables
house.df <- household |> 
  filter(income_inkind != 0,
         Q229NETINCOME != 0,
         expenditure_inkind !=0) |>
  dplyr::select(
    UQNO, income_inkind, expenditure_inkind, Q229NETINCOME, 
    SETTLEMENT_TYPE, hholds_wgt, province_code, SexOfHead) |>
  transmute(SexOfHead,
    UQNO = as.character(UQNO), 
    psu = substr(UQNO,1,11),
    du = substr(UQNO,12,16),
    hhnum = substr(UQNO,17,18),
    hhwt = hholds_wgt,
    lny = log(as.numeric(income_inkind)/12),
    lnminq = log(as.numeric(Q229NETINCOME)),
    lnx = log(as.numeric(expenditure_inkind/12)),
    province = as_factor(province_code),
    settle = as_factor(SETTLEMENT_TYPE)
    ) |>
  droplevels() 
  


#merge above 4 dataframes to get a dataframe LCS2014
df <- person1 |>
  full_join(house.df, by="UQNO") |>
  drop_na() 

LCS2014 <- df |>
  filter(adults == 2,
         kids <= 1) 

# here is the data for the lab
#lab4.df <- df
#save(lab4.df, file = "../../../Data/lab4.Rdata")
```

## Manipulate it (for lasso - not used today)

```{r}
#| label: lcs2
#| dependson: lcs

house.df2 <- household |>
  dplyr::select(
    UQNO, income_inkind, Q229NETINCOME, province_code, expenditure_inkind,
    Ageofhead, SexOfHead, PopGroupOfHead, hhsize, hholds_wgt) |>
  transmute(
    UQNO = as.character(UQNO), 
    psu = substr(UQNO,1,11),
    du = substr(UQNO,12,16),
    hhnum = substr(UQNO,17,18),
    hhwt = hholds_wgt,
    x = as.numeric(expenditure_inkind),
    x.ihs = ihs(x/12),
    y = as.numeric(income_inkind),  # 63 observations have 0 values... which(house.df$y==0)
    y.ihs = ihs(y/12),
    min.ihs = ihs(Q229NETINCOME),
    province = as_factor(province_code),
    age = as.integer(Ageofhead),
    male = as_factor(SexOfHead),
    pop.group = as_factor(PopGroupOfHead)
    ) |>
  droplevels()

lasso.df <- person1 |>
  full_join(house.df2, by="UQNO") |>
  drop_na() 

df.x <- model.matrix(object = min.ihs ~ x.ihs + y.ihs + kids + adults 
                     + age + male + province + pop.group,
                     data = lasso.df) 
```


# Variances and Testing

## As we showed last time

- The variance definition

$$
V(\hat{\beta}|X) = E \left[ \left( \hat{\beta} - \beta \right) \left( \hat{\beta} - \beta
  \right)' \bigg| X \right]
$$

- In matrix notation

$$
V(\hat{\beta}| X) = E \Biggl[ [X'X]^{-1}X'uu'X[X'X]^{-1} \Bigg|X \Biggr] 
$$

## How to estimate it?

- Basic analogy principle
- Use an 'estimate' for an actual value
- Thus:

$$
V(\hat{\beta}| X) = E \Biggl[ [X'X]^{-1}X'\hat{u}\hat{u}'X[X'X]^{-1} \Bigg|X \Biggr] 
$$

## Some thoughts

- We do not expect any particular $\hat{u}_i$ to be zero
- Instead, $E[\hat{u}|X] = 0$
- Technically, we are squaring, anyway..
- Take a look at the [sandwich package](https://cran.r-project.org/web/packages/sandwich/index.html)

## Some 'corrections'

1. HC0: use $\hat{u}_i^2$

2. HC1: use $\frac{n}{n-k} \cdot \hat{u}_i^2$

3. HC2: use $\frac{\hat{u}_i^2}{1-h_{ii}}$, where $h_{ii}$ is the term from the predictor matrix discussed last week

4. HC3: use $\frac{\hat{u}_i^2}{(1-h_{ii})^2}$

5. HC4: use $\frac{\hat{u}_i^2}{(1-h_{ii})^{\delta_i}}$, where $\delta_i = \min \{4, h_i/\overline{h} \}$

6. HC1 - HC4 are small sample adjustments

## Clusters

1. Really only for HC2 and HC3

2. HC2: use $\frac{G-1}{G} \left(I_{nG} - H_{GG} \right)^{-1}$, where $nG$ is the number of observations in cluster $G$, and $H_{GG}$ is the prediction matrix for cluster $G$.

3. HC3: use $\frac{G-1}{G} \left(I_{nG} - H_{GG} \right)^{-2}$, where $nG$ is the number of observations in cluster $G$, and $H_{GG}$ is the prediction matrix for cluster $G$.

4. There are further corrections for multiple cluster dimensions, which we might see in panels, for instance.

5. Bootstrapping is also an option..

## Estimation plan (example)

- Regress log minimum income on actual income, number of adults in household, and settlement type
- Present OLS, HC2 and HC2 with clusters.
- See next
- What do we see?

## Code

```{r}
#| eval: false

model <- lm(lnminq ~ lny + adults + settle, 
            data = df)

robust_se <- sqrt(diag(vcovHC(model, type = "HC2")))

cluster_se <- sqrt(diag(vcovCL(model, cluster = ~ psu, type = "HC2")))

stargazer(model, model, model, 
          type = "html",
          se = list(NULL, robust_se, cluster_se),
          title = "Regression Results with Different Standard Errors",
          dep.var.labels = "Log minimum income",
          keep = c("lny", "adults"),
          model.numbers = FALSE,
          column.labels = c("Homoskedastic","HC2","Cluster HC2"),
          no.space = TRUE, 
          keep.stat = "n", 
          header = FALSE, 
          digits = 5)

```


## Regression Results with Different Standard Errors

```{r}
#| label: simple-reg
#| dependson: lcs
#| echo: false
#| results: asis

model <- lm(lnminq ~ lny + adults + settle, 
            data = df)

robust_se <- sqrt(diag(vcovHC(model, type = "HC2")))

cluster_se <- sqrt(diag(vcovCL(model, cluster = ~ psu, type = "HC2")))

stargazer(model, model, model, 
          type = "html",
          se = list(NULL, robust_se, cluster_se),
          dep.var.labels = "Log minimum income",
          keep = c("lny", "adults"),
          model.numbers = FALSE,
          column.labels = c("Homoskedastic","HC2","Cluster HC2"),
          no.space = TRUE, 
          keep.stat = "n", 
          header = FALSE, 
          digits = 5)

```


# Testing: Single Parameter

## Back to the variance: a quadratic form


$$
\begin{split}
\left( \hat{\beta}-\beta \right) \left( \hat{\beta}-\beta \right)' &= 
  \left( \left(X'X\right)^{-1}X'u \right) \left( \left(X'X\right)^{-1}X'u \right)' \\
  &= \left(X'X\right)^{-1}X'uu'X\left(X'X\right)^{-1} \\
  &=  A^{-1} B A^{-1}
\end{split}
$$

+ So, let us look at $A$ and $B$


## Pieces of the QF: $A$

+ Start with $A$, which is from observed data

$$
\begin{split}
A &= X'X \\
\hat{A} &= N^{-1} \left( \sum_{i=1}^N x_i'x_i \right)
\end{split}
$$

+ Simple average
+ Built-up from the observation level

## Pieces of the QF: $B$

+ Let us look at $B$ (has unobserved information)

\begin{equation}\begin{split}
B &= X'uu'X \\
\hat{B} &= N^{-1} \left( \sum_{i=1}^N x_i'\hat{u}_i^2 x_i \right) \\
 &= N^{-1} \left( \sum_{i=1}^N \hat{u}_i^2x_i'x_i \right) 
\end{split}\end{equation}

## Distribution of estimators I

+ Substituting sample means for population means

$$
\begin{aligned}
\hat{\beta} &\xrightarrow{p} \left( \frac{X'X}{N} \right)^{-1} \left(
  \frac{X'y}{N} \right) = \beta \\
\widehat{V \left[ \hat{\beta} \right] } & \xrightarrow{p} 
 \left( \frac{X'X}{N} \right)^{-1}  \left( \frac{ X'uu'X}{N}
  \right)  \left(\frac{X'X}{N} \right)^{-1}   \\
  &= N^{-1} A^{-1}BA^{-1}
\end{aligned}
$$

## Distribution of estimators II

+ Taking advantage of one of the many CLTs

$$
\begin{split}
\hat{\beta} &\xrightarrow{d} \mathcal{N} \left( \beta , N^{-1}\hat{A}^{-1}\hat{B}\hat{A}^{-1}
\right) \\
\hat{\beta} - \beta &\xrightarrow{d} \mathcal{N} \left( 0 , N^{-1}\hat{A}^{-1}\hat{B}\hat{A}^{-1}
\right) \\
\sqrt{N} \left(\hat{\beta} - \beta \right)  &\xrightarrow{d} \mathcal{N} \left( 0 , \hat{A}^{-1}\hat{B}\hat{A}^{-1}
\right) \\
\end{split}
$$

## Heteroscedasticity-robust test statistics

+ Pull out $t-$ statistics or $z-$statistics (using $z$ for simplicity)
+ Replace $A^{-1}BA^{-1}$ with $\hat\Omega$

$$
\frac{\hat{\beta}_j - \beta_j}{ \sqrt{\hat\Omega[j,j] /N} }
\xrightarrow{d} \mathcal{N} (0,1)
$$




# Testing: Multiple Parameter

## An example setup

+ Cobb-Douglas production model 

$$
y=AK^{\beta_0} L^{\beta_1} E^{\beta_2}
$$

+ OLS in log

$$
\ln y= \ln A + \beta_0 \ln K + \beta_1 \ln L + \beta_2 \ln E + u
$$

## A common assumption

+ Constant returns to scale?

$$
\beta_0 + \beta_1 + \beta_2 = 1
$$

+ A linear restriction 
+ On multiple parameters

## Imposing restrictions under homoscedasticity

+ Homoscedasticity leads to an $F-$test
+ General restriction is $R\beta=r$

$$
\begin{pmatrix} 1 & 1 & 1 \end{pmatrix}
\begin{pmatrix} \beta_0 \\ \beta_1 \\ \beta_2 \end{pmatrix} = 
\begin{pmatrix} 1 \end{pmatrix}
$$

## Apply variance rules

+ Under Homoscedasticity

$$
\tilde{\beta} \xrightarrow{d} \mathcal{N} \left( \beta, \sigma^2
  \left( X'X \right)^{-1} \right)
$$

+ The distribution of $R\tilde\beta$

$$
R\tilde{\beta} \xrightarrow{d} \mathcal{N} \left( R \beta, \sigma^2
  R \left( X'X \right)^{-1} R'\right)
$$

## Adjusting the variances

+ Squared standard normal is distributed as a $\chi^2_{J}$
+ $J$ is number of restrictions or "$=$" signs...

$$
Q_1 = \frac{ \left( R\tilde{\beta} - R\beta \right)' \left[R \left(
      X'X \right)^{-1} R' \right]^{-1} \left( R\tilde{\beta} - R\beta
  \right)}{\sigma^2} \sim \chi^2_J
$$

+ By similar reasoning, 

$$
Q_2 = \frac{ (N-k) \hat{\sigma}^2 } {\sigma^2} \sim \chi^2_{N-k}
$$

## Developing the $F$

+ Ratio of two $\chi^2$ random variables, normalized by their degrees of freedom
+ Yields an $F-$statistic ($\sigma^2$ terms cancel, so not included)

$$
\lambda = \frac{ \left( R\tilde{\beta} - R\beta \right)' \left[R \left(
      X'X \right)^{-1} R' \right]^{-1} \left( R\tilde{\beta} - R\beta
  \right)}{ \hat{\sigma}^2} \cdot \frac{N-k}{J} \sim F_{J,N-k}
$$


## Impose the restriction

$$
\lambda = \frac{ \left( R\tilde{\beta} - r \right)' \left[R \left(
      X'X \right)^{-1} R' \right]^{-1} \left( R\tilde{\beta} - r
  \right)}{ \hat{\sigma}^2} \cdot \frac{N-k}{J} \sim F_{J,N-k}
$$

+ After some manipulation

$$
\left( R\tilde{\beta} - r \right)' \left[R \left(
      X'X \right)^{-1} R' \right]^{-1} \left( R\tilde{\beta} - r
  \right) = \text{ESS}_U - \text{ESS}_R
$$

## The test statistic you might remember?

+ The statistic

$$
\frac{ \left(R^2_U - R^2_R \right) /J}
{\left( 1-R^2_U \right)/(N-k)} \sim F_{J,N-k}
$$

+ What does it 'say'??

## Heteroscedasticity effects

+ Heteroskedasticity invalidates the standard $F$
    + A heteroskedasticity robust version exists
    + Here, I describe the Wald version (instead)
+ Define a general linear hypothesis $R\beta=r$
+ This is as before

## Return to the distribution of the estimator


$$
\begin{split}
\hat{\beta} &\xrightarrow{d} \mathcal{N} \left( \beta , N\Omega \right) \\
R \hat{\beta} &\xrightarrow{d} \mathcal{N} \left( R \beta , NR\Omega R' \right) \\
R \hat{\beta} -r &\xrightarrow{d} \mathcal{N} \left( R \beta -r , NR\Omega R' \right) \\
R \hat{\beta} -r &\xrightarrow{d} \mathcal{N} \left( 0 , NR\Omega R' \right) \\
\end{split}
$$
 
+ From this distribution, we can develop a $\chi^2$ distribution

## Needed pieces

+ Definitions:
    + Standard normal deviate: A random sample from standard normal distribution
    + $\chi^2$ Distribution: Sum of squared normal deviates
    + $\chi^2$ Degrees of Freedom: The number of standard normal deviates being summed
+ Above deviate is not standard normal
+ How do we do this in one dimension (Recall from early stats)? 

## Extending


+ How does that translate to multiple dimensions?
+ Taking from above
    + $R\hat{\beta} - r$ is zero mean normal variate
    + We need to divide by square root of variance
    + Requires matrix decomposition, let us skip

## With this in mind...

$$
k^{-1}(R \hat{\beta} -r)' (R\Omega R')^{-1} (R \hat{\beta} -r)  \xrightarrow{d} \chi^2_k 
$$

+ 'Divided' the squared normal deviate by its variance $(R\Omega R')^{-1}$
+ Equivalent to squaring a standard normal deviate
+ This is a Wald test, rather than an $F$

## Test Estimate

- We use the same model from before, but test all the parameters on settlement.
- The test statistics will be for homoskedasticity, thus an F, as well as both heteroskedastic and cluster-heteroskedastic models

## Some code

```{r}
#| eval: false
#| echo: true

# I use the same covariances matrixes.
# Create R matrix
R <- matrix(c(0, 0, 0, 1, 0, 0, #b4 = 0
              0, 0, 0, 0, 1, 0, #b5 = 0
              0, 0, 0, 0, 0, 1), #b6 = 0
            byrow = T, nrow = 3)

robust_vcov <- vcovHC(model, type = "HC2")

cluster_vcov <- vcovCL(model, cluster = ~ psu, type = "HC2")

mtest.F <- linearHypothesis(model, hypothesis.matrix = R, test = "F")
mtest.ch.het <- linearHypothesis(model, 
                                 hypothesis.matrix = R, 
                                 vcov. = robust_vcov,
                                 test = "Chisq")
mtest.ch.hetcl <- linearHypothesis(model, 
                                   hypothesis.matrix = R, 
                                   vcov. = cluster_vcov,
                                   test = "Chisq")

results.tab1 <- cbind("F",mtest.F[2,3], mtest.F[2,5], mtest.F[2,6])
results.tab2 <- cbind("Het",mtest.ch.het[2,2], mtest.ch.het[2,3], 
                      mtest.ch.het[2,4])
results.tab3 <- cbind("ClustHet",mtest.ch.hetcl[2,2], mtest.ch.hetcl[2,3],
                      mtest.ch.hetcl[2,4])

results.tab <- rbind(results.tab1,results.tab2,results.tab3)

kable(results.tab,
      format = "html",
      caption = "F and Chi Sq tests",
      col.names = c("Var-Assump","DF","F or ChiSq", "Prob"))
```


## Some results

```{r}
#| label: multiple-tests
#| dependson: simplereg
#| results: asis
#| echo: false

# I use the same covariances matrixes.
# Create R matrix
R <- matrix(c(0, 0, 0, 1, 0, 0, #b4 = 0
              0, 0, 0, 0, 1, 0, #b5 = 0
              0, 0, 0, 0, 0, 1), #b6 = 0
            byrow = T, nrow = 3)

robust_vcov <- vcovHC(model, type = "HC2")

cluster_vcov <- vcovCL(model, cluster = ~ psu, type = "HC2")

mtest.F <- linearHypothesis(model, hypothesis.matrix = R, test = "F")
mtest.ch.het <- linearHypothesis(model, 
                                 hypothesis.matrix = R, 
                                 vcov. = robust_vcov,
                                 test = "Chisq")
mtest.ch.hetcl <- linearHypothesis(model, 
                                   hypothesis.matrix = R, 
                                   vcov. = cluster_vcov,
                                   test = "Chisq")

results.tab1 <- cbind("F",mtest.F[2,3], mtest.F[2,5], mtest.F[2,6])
results.tab2 <- cbind("Het",mtest.ch.het[2,2], mtest.ch.het[2,3], 
                      mtest.ch.het[2,4])
results.tab3 <- cbind("ClustHet",mtest.ch.hetcl[2,2], mtest.ch.hetcl[2,3],
                      mtest.ch.hetcl[2,4])

results.tab <- rbind(results.tab1,results.tab2,results.tab3)

kable(results.tab,
      format = "html",
      caption = "F and Chi Sq tests",
      col.names = c("Var-Assump","DF","F or ChiSq", "Prob"))
```


# Decomposition

## Counterfactual type questions

+ How might female outcomes (wages, e.g.) change if they were men?
+ We cannot make them men, but we can give them the data characteristics of men
+ Decomposition is one way to examine this


## Atheoretical construct

+ Define a binary "treatment"
+ Assume it is exogenous

$$
D_i = \begin{cases} 1, &\text{If male} \\
  0, & \text{If female}
\end{cases}
$$

+ Define "treatment" potential outcomes:


$$
Y_i = \begin{cases} Y_{1i}, &\text{Outcome, if male} \\
Y_{0i}, & \text{Outcome, if female}
\end{cases} 
$$


## Some formality

+ Condition $Y$ on other variables
+ Separate regressions for both groups

$$
\begin{aligned}
Y_{0i} &= x_{1i} \beta_0 + u_{0i} \\
Y_{1i} &= x_{0i} \beta_1 + u_{1i}
\end{aligned}
$$

+ Responses might differ
+ Characteristics might differ
+ Ignored for now: Structure could differ


## Decomposition foundation


+ Define our estimate of interest

$$
\hat{\Delta} = \overline{Y}_{1i} - \overline{Y}_{0i}
$$

+ Rewrite a previous useful assumption
+ $E[D_i u_{0i}] = E[D_i u_{1i}] = 0$
+ Why this assumption?
+ Eliminates unobservable effects

## Thus...

$$
\hat{\Delta} = \overline{x}_0 \left( \hat{\beta}_1 - \hat{\beta}_0 \right) 
             + \hat{\beta}_1 \left( \overline{x}_1 - \overline{x}_0 \right) 
$$


## Derivation...

$$
\begin{split}
\hat\Delta_2 =& \overline{x}_1 \hat{\beta}_1 - \overline{x}_0 \hat{\beta}_0 
         + \overline{x}_0 \hat{\beta}_1 - \overline{x}_0 \hat{\beta}_1 \\
         =& \overline{x}_0\left( \hat{\beta}_1 - \hat{\beta}_0 \right) 
          + \hat{\beta}_1 \left( \overline{x}_1 - \overline{x}_0 \right) 
\end{split}
$$

+ "2" for twofold decomposition
+ It includes a $\beta$ component: unexplained
+ It includes an $x$ component: explained
+ What is meant by explained and unexplained?

# Extending Decomposition

## Basic setup, repeated

+ Recall the basic model

$$
\begin{aligned}
Y_{0i} &= x_{0i} \beta_0 + u_{0i} \\
Y_{1i} &= x_{1i} \beta_1 + u_{1i}
\end{aligned}
$$

+ We want to look at


$$
\hat{\Delta} = \overline{Y}_1 - \overline{Y}_0 = \overline{x}_1 \hat{\beta}_1 - \overline{x}_0 \hat{\beta}_0
$$

## Including an interaction term

+ Derivation

$$
\begin{split}
\Delta_3 =& \overline{x}_1 \hat{\beta}_1 - \overline{x}_0 \hat{\beta}_0 
          + \left( \overline{x}_0 \hat{\beta}_0 - \overline{x}_0 \hat{\beta}_0 \right)
          + \left( \overline{x}_1 \hat{\beta}_0 - \overline{x}_1 \hat{\beta}_0 \right)
          + \left( \overline{x}_0 \hat{\beta}_1 - \overline{x}_0 \hat{\beta}_1 \right) \\
         =& \overline{x}_0\left( \hat{\beta}_1 - \hat{\beta}_0 \right) 
          + \hat{\beta}_0 \left( \overline{x}_1 - \overline{x}_0 \right) 
          + \hat{\beta}_1 \left( \overline{x}_1 - \overline{x}_0 \right)
          - \hat{\beta}_0  \left( \overline{x}_1 - \overline{x}_0 \right) \\
         =& \overline{x}_0\left( \hat{\beta}_1 - \hat{\beta}_0 \right) 
          + \hat{\beta}_0 \left( \overline{x}_1 - \overline{x}_0 \right) 
          + \left( \hat{\beta}_1 - \hat{\beta}_0 \right) \left( \overline{x}_1 - \overline{x}_0 \right)
\end{split}
$$

+ "3" for threefold decomposition

## Interaction term discussion

+ It still includes:
    + unexplained
    + explained
+ It also includes interaction between coefficients and characteristics
+ Hmmm, what does that measure/represent?
+ Cross-group differences in explanatory variables and coefficients can occur at the same time

# Some Decomp Issues

## What is the right base group?

+ Our previous version...

$$
\begin{split}
\hat\Delta_2 =& \overline{x}_1 \hat{\beta}_1 - \overline{x}_0 \hat{\beta}_0 
         + \overline{x}_0 \hat{\beta}_1 - \overline{x}_0 \hat{\beta}_1 \\
         =& \overline{x}_0\left( \hat{\beta}_1 - \hat{\beta}_0 \right) 
          + \hat{\beta}_1 \left( \overline{x}_1 - \overline{x}_0 \right) 
\end{split}
$$

+ What if we reversed the difference?

$$
\begin{split}
\hat\Delta_2 =& \overline{x}_0 \hat{\beta}_0 - \overline{x}_1 \hat{\beta}_1 
         + \overline{x}_1 \hat{\beta}_0 - \overline{x}_1 \hat{\beta}_0 \\
         =& \overline{x}_1\left( \hat{\beta}_0 - \hat{\beta}_1 \right) 
          + \hat{\beta}_0 \left( \overline{x}_0 - \overline{x}_1 \right) 
\end{split}
$$

## Questions

+ Are these equal?
+ When might they be?
+ Researchers have suggested many "averages" to address this

## What happens with dummies/categories in a decomposition?

+ Look at a simplistic regression
+ Here, $d$ has three levels, one of which is left out
+ For completeness, $d_j = \mathbf{1}(D=j)$, $j=\{1,2,3\}$

$$ 
y_i = \beta_0 + \beta_1 d_{i1} + \beta_2 d_{2i} + e_i
$$

+ As with the choice of base for the decomposition regression
+ The base of the categories affects the "math"

## Add back the left-out dummy?

+ Consider this version of the model

$$ 
y_i = \beta_0 + \beta_1 d_{i1} + \beta_2 d_{2i} + \beta_3 d_{3i} + e_i
$$

+ That would give a DV trap
+ But, can "restrict" $\beta_3 = 0$
+ Define: $a = \sum_{j=1}^3 \beta_j /J$, here, $J=3$

## Relation between $a$ and the $\beta$s?

+ With $a$ above and $(j \neq 0)$
    + $\tilde\beta_0 = \beta_0 + a$
    + $\tilde\beta_j = \beta_j - a$
+ The above model is equivalent to:

$$ 
y_i = \tilde\beta_0 + \tilde\beta_1 d_{i1} + \tilde\beta_2 d_{2i} + \tilde\beta_3 d_{3i} + e_i
$$

+ Subject to the restriction: $\sum_j \tilde{\beta} = 0$

# Brief Aside on Contrasts

## Dummy coding in R 

+ "Dummy" is how economists normally use categories
+ For example, ```factor.var <- factor(var, levels = , labels =)```
    + Levels can be used to create a baseline: first is the base
    + Labels match the order of the levels

## Releveling

+ So, could redefine the factor, ```relevel``` the factor or
+ Can also use ```contr.treatment()``` to change the baseline
    + For example ```factor.var <- contr.treatment(3, base = 2)```
    + This uses group 2 as the baseline of the three levels

## Standard dummy contrasts

+ Under our 'dummy' approach, the intercept is the left-out mean
+ This is a type of "contrast"; for ```contr.treatment(3)```

```{r, results="markdown"}

contr.treatment(3)
```

+ Group 1 is base, and represents intercept
+ Column 2 compares variable 2 to base
+ Column 3 compares variable 3 to base

## Sum Effects coding in R

+ One could consider a "sum"
+ It compares mean of a dependent variable for a given level to the overall mean of the dependent variable ```contr.sum(3)```

```{r}
#| results: markup

contr.sum(3)
```

## Interpret

+ First column compares variable 1 to the grand mean 
+ Second compares variable 2 to the grand mean
+ Grand mean here is the mean of the means of the dependent variable across levels

## Helmert effects in R

+ Helmert coding compares each level of a categorical variable to the mean of subsequent levels of the variable
+ ```myhelmert2 <- matrix(c(2/3, -1/3, -1/3, 0, .5, -.5), ncol = 2)```
+ ```myhelmert2```

```{r}
#| results: markup
myhelmert2 <- matrix(c(2/3, -1/3, -1/3, 0, .5, -.5), ncol = 2)
myhelmert2

```

## Interpreting

+ First  coefficient is mean of level 1 compared to the mean of levels 2, and 3
+ Second  coefficient is mean of level 2 compared to the mean of level 3  

## Reverse helmert effects in R

+ Reverse helmert coding compares each level of a categorical variable to the mean of previous levels of the variable
+ ```contr.helmert(3)```

```{r}
#| results: markup
contr.helmert(3)
```

## Interpreting

+ First regression coefficient is the mean of level 2 compared to the mean of level 1 
+ Second regression coefficient is mean of level 3 minus the mean of levels 1 and 2
+ Looking "vertically"


# Back to Decomp Example


## A simple decompsition

+ Model
    + Dep Var: Minimum required income (inverse hyperbolic sine)
    + Ind Var: Household income (inverse hyperbolic sine)
    + Decomp Group: Households with 1 child v Households with 0

## The "Question" 

+ Is there a difference in minimum required income, due to having a child in the household?

+ Is that difference related to differences in total household income (explained) or differences in the relationship between income and minimum income across groups (unexplained)?

## Estimates code


```{r}
#| eval: false

# Note this bootstrap does not necessarily correct for heteroskedasticity
kids.dec <- oaxaca(lnminq ~ lny | kids, 
                   data = LCS2014,
                   group.weights = 0.5,
                   R = 399)

kable(t(kids.dec$threefold$variables),
      format = "html",
       caption = "Three-fold decomp")
```


## Decomposition on first child: estimates

```{r}
#| label: decompres
#| dependson: lcs
#| results: asis
#| echo: false

# Note this bootstrap does not necessarily correct for heteroskedasticity
kids.dec <- oaxaca(lnminq ~ lny | kids, 
                   data = LCS2014,
                   group.weights = 0.5,
                   R = 399)

kable(t(kids.dec$threefold$variables),
      format = "html",
       caption = "Three-fold decomp")
```

## Answers

+ There does not seem to be much difference
+ Does not seem to be explained or unexplained...
+ Almost $\ln y$ statistical significance, but small

## Endowments plot code

```{r}
#| eval: false

dec.name <- names(kids.dec$x$x.mean.A) # gets names of variables

plot(kids.dec, components="endowments")
```


## Decomposition on first child: plot endowments effect

```{r}
#| label: decomp-plot
#| dependson: decompres
#| fig-cap: "Decomposition results - endowments"
#| echo: false

dec.name <- names(kids.dec$x$x.mean.A) # gets names of variables

plot(kids.dec, components="endowments")
```

## Coefficients plot code


```{r}
#| eval: false

plot(kids.dec, components="coefficients")
```    


## Decomposition on first child: plot coefficients effect

```{r}
#| label: decomp-plot2
#| dependson: decompres
#| fig-cap: "Decomposition results - coefficients"
#| echo: false

plot(kids.dec, components="coefficients")
```    

## Decomposition on first child: plot interactions effect

```{r}
#| label: decomp-plot3
#| dependson: decompres
#| fig-cap: "Decomposition results - interactions"
#| echo: false

plot(kids.dec, components="interaction")
```    

# References

## References
